\chapter{Introduction}
\label{ch:introduction}

% ─────────────────────────────────────────────────────────────────────────────
\section{Background and Motivation}
\label{sec:intro_background}
% ─────────────────────────────────────────────────────────────────────────────

The educational technology sector has undergone significant expansion over the
past decade, yet the dominant paradigm in digital learning platforms remains
one-size-fits-all content delivery. A learner who already understands tectonic
plate mechanics receives the same sequence of questions as one who has never
encountered the concept; a learner who has mastered map reading is presented
with the same map skills introduction as a complete novice. This homogeneity
is not a technical constraint but a design choice --- and one that the
empirical literature consistently identifies as suboptimal.

\textcite{bloom1984} demonstrated that one-to-one human tutoring, which
adapts to the learner's current ability in real time, produces learning
outcomes approximately two standard deviations above conventional classroom
instruction. \textcite{vanlehn2011} subsequently showed that step-based
Intelligent Tutoring Systems (ITS) can approach this performance at scale,
achieving effect sizes statistically indistinguishable from human tutoring.
The implication is clear: the adaptive feedback loop is the critical variable,
not the delivery medium or the domain.

Despite this evidence, adaptive ITS remain largely confined to mathematics and
programming domains, where knowledge structures are well-defined and assessment
is unambiguous. Geography --- a subject that integrates declarative factual
knowledge, procedural reasoning, and conceptual understanding --- has received
almost no attention in the adaptive learning literature \parencite{bednarz2013}.
The UK KS3/GCSE Geography curriculum, with its distinctive structure spanning
physical geography, human geography, and geographical skills, represents both
an underserved educational context and a meaningful test case for ITS methods
applied to non-STEM knowledge domains.

% ─────────────────────────────────────────────────────────────────────────────
\section{Research Question}
\label{sec:intro_rq}
% ─────────────────────────────────────────────────────────────────────────────

This project addresses the following research question:

\begin{quote}
\textit{Can a hybrid IRT 3PL + BKT adaptive engine, implemented in a
production web application, deliver personalised question sequencing that
produces measurable within-session learning gains for learners studying
UK KS3/GCSE Geography content?}
\end{quote}

The question has three components: an \textit{algorithmic claim} (the hybrid
IRT+BKT engine can perform valid adaptive selection), a \textit{technical
claim} (this can be implemented in a production-deployed web application), and
an \textit{empirical claim} (the system produces measurable learning gains).
The evaluation strategy in Chapter~\ref{ch:results} addresses all three,
with the acknowledgement that the empirical claim is evaluated at
proof-of-concept scale rather than at statistical significance.

% ─────────────────────────────────────────────────────────────────────────────
\section{Project Overview}
\label{sec:intro_overview}
% ─────────────────────────────────────────────────────────────────────────────

GeoMentor is an adaptive intelligent tutoring system for UK KS3/GCSE
Geography, accessible at \url{https://e-learning-final-project.vercel.app}.
The system implements a hybrid IRT 3PL + BKT algorithm that continuously
estimates learner ability and Knowledge Component mastery, selects questions
targeting the learner's Zone of Proximal Development, and enforces Bloom's
taxonomy cognitive prerequisites across a 13-KC curriculum graph.

The system was developed as the artefact for this Honours Stage Project over
ten two-week sprints from October 2025 to May 2026, following a Design Science
Research methodology \parencite{hevner2004}. It is production-deployed,
publicly accessible, and open-source, providing a proof-of-concept
implementation that can be independently verified, reproduced, and extended.

The scope was deliberately concentrated on a single domain following a Sprint 5
risk assessment that identified multi-domain calibration as infeasible within
the available timeline. The system's architecture is domain-agnostic, and
Cross-Domain Knowledge Transfer is identified as the primary architectural
stretch goal for future development (Section~\ref{sec:lr_future}).

% ─────────────────────────────────────────────────────────────────────────────
\section{Contributions}
\label{sec:intro_contributions}
% ─────────────────────────────────────────────────────────────────────────────

This project makes the following contributions:

\begin{enumerate}[leftmargin=2em]

  \item \textbf{A production-deployed hybrid IRT 3PL + BKT adaptive engine
  for UK Geography.} The first publicly accessible ITS targeting the UK
  KS3/GCSE Geography curriculum, extending the empirical evidence base for
  adaptive learning beyond traditional STEM domains.

  \item \textbf{A credit-factor modification to the standard BKT update.}
  A novel modification of the BKT learning transition equation that weights
  correct responses by a graduated credit factor reflecting the maximum hint
  level accessed, formalising the evidential relationship between scaffolding
  usage and mastery evidence.

  \item \textbf{Bloom-gated adaptive progression.} An implementation of
  Bloom's taxonomy as a computational selection constraint, enforcing that
  learners consolidate recall-level knowledge before encountering
  application-level items --- a feature absent from standard IRT-only CAT
  implementations.

  \item \textbf{A five-track multi-method evaluation framework for small-sample
  ITS validation.} A structured evaluation approach combining algorithmic
  validation, engagement data, usability metrics, misconception analysis, and
  robustness monitoring, designed for contexts where inferential statistics are
  not viable due to small sample sizes.

  \item \textbf{An open-source domain-agnostic ITS architecture.} The codebase
  provides a reusable template for researchers and developers building adaptive
  systems for non-STEM declarative-knowledge domains, with the KC graph,
  question bank, and IRT parameter store fully parameterised and independent
  of Geography-specific logic.

\end{enumerate}

% ─────────────────────────────────────────────────────────────────────────────
\section{Scope and Delimitations}
\label{sec:intro_scope}
% ─────────────────────────────────────────────────────────────────────────────

The following boundaries apply to this project:

\begin{itemize}[leftmargin=2em]
  \item \textbf{Domain.} UK KS3/GCSE Geography only (AQA specification). No
    other subject domain is supported in the current release, though the
    architecture is designed for extension.
  \item \textbf{Assessment type.} Multiple-choice questions with four options.
    Extended writing, diagram annotation, and data interpretation items are
    outside scope.
  \item \textbf{Participant population.} University of Hull students
    (convenience sample). The system targets KS3/GCSE students but was
    evaluated with adults due to ethical and logistical constraints on
    secondary school access.
  \item \textbf{Algorithm.} IRT 3PL + BKT hybrid. Deep Knowledge Tracing,
    Elo, and 2PL models were evaluated and rejected
    (Section~\ref{sec:lr_algorithms}).
  \item \textbf{Evaluation scale.} Proof-of-concept validation at $n = 4$.
    Inferential statistical claims are not made.
\end{itemize}

% ─────────────────────────────────────────────────────────────────────────────
\section{Ethical and Professional Considerations}
\label{sec:intro_ethics}
% ─────────────────────────────────────────────────────────────────────────────

All participant data was collected under informed written consent. No
personally identifiable data beyond email address and name (required for
authentication) was collected; session performance data is stored under
a pseudonymous user ID and is accessible only to the authenticated user and
to the researcher under the study protocol.

Artificial intelligence tools were used in the development and writing of this
project. Specifically: authentication boilerplate, UI scaffolding, and Vercel
deployment configuration received AI assistance. The research design, KC
ontology, IRT calibration strategy, BKT credit-factor modification, and the
adaptive engine's core algorithm were designed and implemented by the author.
Full details of AI usage are provided in the AI Declaration appendix, in
compliance with the University of Hull's Academic Integrity Policy.

% ─────────────────────────────────────────────────────────────────────────────
\section{Report Structure}
\label{sec:intro_structure}
% ─────────────────────────────────────────────────────────────────────────────

The remainder of this report is organised as follows.

Chapter~\ref{ch:literature_review} presents a critical review of the
theoretical and empirical foundations of adaptive intelligent tutoring systems,
covering pedagogical theory, adaptive algorithms, Computerised Adaptive Testing
principles, empirical effectiveness evidence, implementation challenges, and
ethical considerations. Forty-five sources are reviewed and five research gaps
are identified.

Chapter~\ref{ch:methodology} describes the research design, KC ontology,
adaptive algorithm design with full mathematical derivation, IRT calibration
pipeline, pre-test seeding, scaffolding design, evaluation protocol, agile
development methodology, project Gantt chart, risk register, and an honest
account of what was and was not delivered.

Chapter~\ref{ch:implementation} describes the technical implementation: the
database schema, adaptive engine code, API route design, security
implementation, frontend components, and challenges encountered. Deviations
from the methodology specification are documented with justifications.

Chapter~\ref{ch:results} presents the five-track multi-method evaluation:
algorithm validation, engagement data, SUS usability scores, misconception
analysis, and system robustness metrics.

Chapter~\ref{ch:discussion} interprets the results, compares the system with
related work, defends the scope reduction as a research strength, analyses
limitations and threats to validity, documents what could not be delivered,
identifies future work, and reflects on CSSE competency demonstration.

Chapter~\ref{ch:conclusion} summarises contributions, revisits the research
question, situates the project's broader significance, and offers a final
statement.
