\chapter{Literature Review}
\label{ch:literature_review}

% ─────────────────────────────────────────────────────────────────────────────
\section{Introduction}
\label{sec:lr_intro}
% ─────────────────────────────────────────────────────────────────────────────

Traditional e-learning platforms predominantly deliver static, one-size-fits-all
content despite substantial empirical evidence that personalised instruction
improves knowledge retention, engagement, and long-term learning outcomes
\parencite{vanlehn2011, matos2025, pane2017}. This chapter critically examines
the theoretical and empirical foundations of adaptive intelligent tutoring
systems (ITS), with particular focus on the algorithmic approaches, evaluation
methodologies, and ethical considerations relevant to the system developed in
this project.

The review is organised as follows. Section~\ref{sec:lr_theory} examines the
pedagogical theories underpinning adaptive learning, including Vygotsky's Zone
of Proximal Development, Bloom's mastery learning framework, cognitive load
theory, and spaced retrieval practice. Section~\ref{sec:lr_algorithms} provides
a critical comparative analysis of the primary adaptive algorithms: Elo rating,
Item Response Theory (IRT), Bayesian Knowledge Tracing (BKT), Deep Knowledge
Tracing, and the hybrid IRT+BKT approach adopted in this project.
Section~\ref{sec:lr_cat} examines Computerised Adaptive Testing principles,
including item selection criteria and ability estimation methods.
Section~\ref{sec:lr_effectiveness} synthesises empirical evidence of ITS
effectiveness across domains. Section~\ref{sec:lr_challenges} addresses
implementation challenges including the cold-start problem, IRT calibration with
small samples, and usability. Section~\ref{sec:lr_ethics} considers ethical
dimensions: transparency, data protection, and algorithmic bias.
Section~\ref{sec:lr_gaps} identifies the research gaps this project addresses,
and Section~\ref{sec:lr_future} discusses future directions including
cross-domain knowledge transfer. Section~\ref{sec:lr_conclusion} synthesises
the review's findings into design implications.

% ─────────────────────────────────────────────────────────────────────────────
\section{Theoretical Foundations of Adaptive Learning}
\label{sec:lr_theory}
% ─────────────────────────────────────────────────────────────────────────────

\subsection{The 2-Sigma Problem and the Promise of ITS}
\label{subsec:lr_2sigma}

The empirical case for adaptive personalised instruction originates with
\textcite{bloom1984}, whose experimental work demonstrated that one-to-one
human tutoring produces learning outcomes approximately two standard deviations
above conventional classroom instruction (effect size $d = 2.0$). Bloom termed
this the ``2-sigma problem'': individual tutoring is categorically more
effective than group instruction, yet remains economically unfeasible at scale.
This finding established the central design challenge for educational technology
research: can scalable automated systems replicate the adaptive feedback
mechanisms of a skilled human tutor?

\textcite{vanlehn2011}, in a landmark meta-analysis of 60 controlled studies,
partially resolves this question. His findings demonstrate that human tutoring
achieves $d = 0.79$ (not $d = 2.0$, reconsidering Bloom's original conditions),
and that step-based ITS (those providing feedback on individual solution steps
rather than only final answers) achieve $d = 0.76$, statistically
indistinguishable from human tutoring. This ``interaction plateau hypothesis''
has profound design implications: the critical determinant of ITS effectiveness
is not algorithmic sophistication but interaction granularity. Systems providing
step-level adaptive feedback can approach human tutoring effectiveness regardless
of whether the underlying algorithm is Bayesian, logistic, or rule-based.

\textcite{chi2008} provide a complementary account: the effectiveness of human
tutors derives less from domain expertise and more from their ability to elicit
and respond to student reasoning in real time. This suggests that ITS design
should prioritise responsive feedback over predictive accuracy, a principle
that directly informs the scaffolding architecture of the system presented in
this project (Section~\ref{sec:meth_scaffolding}).

\subsection{Zone of Proximal Development}
\label{subsec:lr_zpd}

Vygotsky's \parencite*{vygotsky1978} sociocultural theory provides the
foundational framework for adaptive question sequencing through the concept of
the Zone of Proximal Development (ZPD). The ZPD defines the critical cognitive
space between what a learner can achieve independently and what they can achieve
with appropriate scaffolding. Crucially, scaffolding must be precisely
calibrated: tasks below the ZPD produce boredom; tasks above it produce
cognitive overload; only tasks within the ZPD produce optimal learning.

The technical operationalisation of the ZPD in an ITS requires a continuous
estimate of learner ability and a mechanism for selecting items whose difficulty
is matched to that estimate. In this system, the ZPD is operationalised through
the IRT target probability $P(\theta) \approx 0.70$: the next question selected
is the item of highest psychometric information at a difficulty level where the
learner has approximately 70\% probability of success. This is neither trivial
(which produces no learning) nor insurmountable (which produces frustration),
implementing the ZPD as a computational targeting criterion.

\textcite{park2004} note that ZPD theory places demands on adaptive systems
beyond simple difficulty sequencing. In Geography specifically, the domain
comprises interconnected declarative knowledge (factual recall of locations and
physical processes), procedural knowledge (applying geographic frameworks to
novel data), and conceptual understanding (explaining spatial patterns and causal
relationships). The prerequisite graph implemented in this system, encoding
structural dependencies between 13 Knowledge Components, is the knowledge
representation mechanism enabling the adaptive engine to navigate this
multi-dimensional knowledge space.

\subsection{Mastery Learning and Bloom's Taxonomy}
\label{subsec:lr_mastery}

\textcite{bloom1984} proposed mastery learning as a pedagogical approach in
which progression through a curriculum is gated by demonstrated competence at
each stage rather than by time-on-task. Students who demonstrate mastery of
prerequisite knowledge are permitted to advance; those who do not receive
targeted remediation until mastery is achieved. Bloom's original framework
predicts that given appropriate instruction and sufficient time, 90\% of
students can achieve what only the top 20\% achieve under conventional
instruction.

The implementation of mastery learning in a digital system requires both a
mastery criterion and a progression gate. \textcite{ritter2016} demonstrate that
mastery thresholds must be precisely calibrated: thresholds above 90\% accuracy
create unnecessary cognitive burden, while thresholds below 60\% permit
knowledge gaps that undermine subsequent learning. Their analysis of
large-scale Cognitive Tutor deployments positions 80\% accuracy as an optimal
threshold. In this system, the BKT mastery gate requires $p(\text{Learned}) \geq
0.80$ per Knowledge Component before the learner can progress to dependent KCs.

Bloom's cognitive taxonomy \parencite{bloom1956}, subsequently revised by
\textcite{anderson2001}, provides the hierarchical framework for question
difficulty beyond simple parametric difficulty. The revised taxonomy identifies
six cognitive levels (Remembering, Understanding, Applying, Analysing,
Evaluating, and Creating), with each level requiring cognitive operations of
greater abstraction than the last. This project implements a three-level
approximation: Bloom Level 1 (Remembering: recall of geographic facts), Level 2
(Understanding: explanation of geographic processes), and Level 3 (Applying:
interpretation of geographic data in novel contexts). The adaptive engine
enforces a Bloom escalation gate: Level 2 items are unlocked when
$p(\text{Learned}) \geq 0.40$ for a KC or when a direct prerequisite KC exceeds
$0.50$; Level 3 items require $p(\text{Learned}) \geq 0.60$. This ensures that
learners master declarative knowledge before encountering analytical demands,
implementing cognitive scaffolding aligned with the taxonomy's hierarchical
structure.

\subsection{Spaced Repetition and Successive Relearning}
\label{subsec:lr_spacing}

A body of evidence spanning over a century establishes that distributed practice,
defined as spacing repetitions of study material over time rather than massing
them in a single session, substantially improves long-term retention
\parencite{cepeda2006}. The spacing effect is one of the most replicated
findings in experimental psychology and has direct implications for adaptive
system design.

\textcite{pavlik2005} formalise the spacing effect through an
activation-based memory model, demonstrating that the optimal inter-repetition
interval depends on the current activation strength of a memory trace. Items
with high activation (recently practised) benefit from longer gaps; items with
low activation (not recently practised, or not yet consolidated) benefit from
shorter gaps and more repetition. Successive Relearning, the practice of
re-presenting material after forgetting has begun rather than before, is
particularly effective for long-term retention, producing superior recall at
delayed tests compared to massed practice or standard spaced repetition
\parencite{kornell2009}.

This project implements Successive Relearning through the BKT decay detection
mechanism: Knowledge Components whose $p(\text{Learned})$ has declined below
the mastery threshold since a prior session are flagged as decayed and receive a
$1.5\times$ information boost in the next-question selection process. This
prioritises re-presentation of consolidating material over introduction of new
content, directly implementing the successive relearning principle in the
adaptive selection criterion.

\subsection{Cognitive Load Theory}
\label{subsec:lr_clt}

\textcite{sweller1988} proposed cognitive load theory (CLT) to account for the
conditions under which instructional design either facilitates or impedes
learning. CLT distinguishes three types of cognitive load: intrinsic load
(determined by the complexity of the material and the learner's schema
development), extraneous load (imposed by poor instructional design), and
germane load (the cognitive effort directed toward schema construction and
automation).

CLT has direct implications for adaptive system design: systems should minimise
extraneous load through clean interface design while maintaining sufficient
intrinsic load to support learning. The graduated hint system in this project
manages intrinsic load by providing incremental scaffolding: conceptual hints at
Level 1 restating the question in simpler terms, procedural hints at Level 2
providing a strategy, and worked examples at Level 3. Each hint level reduces
intrinsic load while also reducing the evidential weight of a subsequent correct
answer \parencite{koedinger2007}. The credit factor modification to BKT formalises
this relationship: a correct answer after a worked example (credit factor 0.50)
contributes less to the mastery estimate than an unaided correct answer (credit
factor 1.00), reflecting the reduced cognitive demand of hint-assisted success.

% ─────────────────────────────────────────────────────────────────────────────
\section{Adaptive Algorithms: A Critical Comparative Analysis}
\label{sec:lr_algorithms}
% ─────────────────────────────────────────────────────────────────────────────

The choice of adaptive algorithm is a critical architectural decision with
long-term consequences for system effectiveness, interpretability, and
maintenance. \textcite{matos2025} identify three dominant approaches in current
ITS deployments: Elo-based rating systems, probabilistic models (IRT and BKT),
and deep learning approaches. Each is evaluated here against four criteria
relevant to this project: \textit{data requirements}, \textit{domain
suitability} (multi-KC, prerequisite-structured), \textit{interpretability},
and \textit{calibration complexity}.

\subsection{Elo Rating System}
\label{subsec:lr_elo}

Originally developed for chess player ranking \parencite{elo1978}, the Elo system
models learner ability and item difficulty on a shared continuous scale, updating
ratings after each interaction via:

\begin{equation}
  R_i^{\text{new}} = R_i + K \cdot \left(\text{outcome} - P(\text{correct})\right)
  \label{eq:elo_update}
\end{equation}

\noindent where $P(\text{correct}) = 1 / (1 + 10^{(R_j - R_i)/400})$, $R_i$ is
the learner ability rating, $R_j$ is the item difficulty rating, and $K$ is the
update step size. \textcite{pelanek2016} demonstrates that Elo systems achieve
approximately 92\% prediction accuracy after only 25--30 learner interactions,
making them attractive for resource-constrained deployments.

However, Elo's simplicity imposes critical limitations for this project's
requirements. First, Elo treats all items as equivalent in discriminating power:
an item that distinguishes ability at a wide range of levels is treated
identically to one that only discriminates near a single difficulty point. In
Geography, items vary considerably in their discriminating properties due to the
range of cognitive demands across the three Bloom levels. Second, Elo operates
as a single global ability estimate: it cannot model a learner who has mastered
Tectonic Hazards but not yet mastered Urban Issues. This per-KC resolution is
essential for a 13-KC curriculum with prerequisite dependencies. Third, Elo
provides no modelling of knowledge decay, making it unsuitable for longitudinal
study designs incorporating 7-day retention assessment. For these reasons, Elo
was evaluated and rejected in favour of the IRT+BKT hybrid.

\subsection{Item Response Theory (3PL)}
\label{subsec:lr_irt}

Item Response Theory \parencite{embretson2000} models the probability of a correct
response as a function of the learner's latent ability $\theta$ and item-specific
parameters. The three-parameter logistic model \parencite{birnbaum1968} is:

\begin{equation}
  P(X=1 \mid \theta) = c + \frac{1-c}{1 + e^{-a(\theta - b)}}
  \label{eq:irt_3pl}
\end{equation}

\noindent where $a$ is item discrimination (steepness of the item characteristic
curve), $b$ is item difficulty (the ability level at which $P = 0.5(1+c)$), and
$c$ is the pseudo-guessing parameter (lower asymptote). The 3PL model
explicitly captures three distinct measurement phenomena that simpler models
cannot: items vary in their discriminating power ($a$), in their difficulty
($b$), and in the probability that a low-ability learner answers correctly by
chance ($c$). The guessing parameter is particularly important for this project:
Geography multiple-choice items with four options carry a non-trivial guessing
probability ($c = 0.25$ for four-option items), and ignoring this would
systematically overestimate learner ability following correct responses.

\subsubsection*{Limitations and Calibration Requirements}

The principal limitation of IRT is its calibration data requirement.
Traditional item parameter estimation via maximum likelihood requires 500+ response
observations per item for stable parameter recovery \parencite{lord1980}. This is
infeasible for a student project with a question bank calibrated from pilot data.
Two mitigations are available: first, fixing the guessing parameter $c$ at
its theoretical value (0.25 for four-option items) rather than estimating it
from data, which substantially stabilises the estimation of $a$ and $b$;
second, adopting the more parsimonious Rasch (1PL) model with fixed discrimination
and no guessing parameter when convergence diagnostics reveal 3PL instability.
This project implements both mitigations: $c = 0.25$ is fixed as an equality
constraint in the calibration pipeline, and a fallback to Rasch is documented
in the risk register (Section~\ref{sec:meth_evaluation}).

IRT has no native mechanism for tracking knowledge acquisition over time; it
estimates ability at a point in time from a response pattern. For multi-KC
curricula requiring longitudinal tracking, IRT must be combined with a
knowledge state model, hence the hybrid architecture.

\subsection{Bayesian Knowledge Tracing}
\label{subsec:lr_bkt}

BKT \parencite{corbett1995} conceptualises learning as a hidden Markov process
with binary latent states: a learner either knows or does not know a Knowledge
Component. Four parameters govern the model per KC:

\begin{align}
  p(L_0) &= \text{initial probability of knowing the KC} \label{eq:bkt_l0}\\
  p(T)   &= \text{probability of learning transition (not-know} \to \text{know)} \label{eq:bkt_t}\\
  p(S)   &= \text{probability of slip (correct response despite not knowing)} \label{eq:bkt_s}\\
  p(G)   &= \text{probability of guess (incorrect response despite knowing)} \label{eq:bkt_g}
\end{align}

After each observation, the posterior probability of knowing the KC is updated
via Bayes' theorem:

\begin{equation}
  p(L_n \mid \text{obs}) = \frac{P(\text{obs} \mid L_n) \cdot p(L_n)}{P(\text{obs} \mid L_n) \cdot p(L_n) + P(\text{obs} \mid \neg L_n) \cdot (1 - p(L_n))}
  \label{eq:bkt_update}
\end{equation}

\noindent where $P(\text{correct} \mid L_n) = 1 - p(S)$ and
$P(\text{correct} \mid \neg L_n) = p(G)$. This is then used to compute the
probability of knowing after a potential learning transition:

\begin{equation}
  p(L_{n+1}) = p(L_n \mid \text{obs}) + \bigl(1 - p(L_n \mid \text{obs})\bigr) \cdot p(T)
  \label{eq:bkt_transition}
\end{equation}

The critical strength of BKT for this project is that it maintains a separate
knowledge state per KC, enabling mastery-gated progression: a learner cannot
advance to UK\_COASTAL until UK\_CLIMATE is mastered, because coastal process
understanding requires prior knowledge of climatic drivers. This per-KC
resolution is unavailable in global ability models such as IRT or Elo.

\textcite{beck2007} identify a parameter identifiability problem in BKT: for
certain parameter combinations, multiple parameter sets produce identical model
predictions, making maximum likelihood estimation unreliable with sparse data.
This is mitigated in this project by setting informed priors based on domain
analysis (Geography KCs at Level 1 difficulty are assigned higher $p(L_0)$
than Level 3 KCs) and by fixing $p(S)$ and $p(G)$ at domain-standard values
following \textcite{pardos2011}.

\textbf{Credit factor modification.} Standard BKT assumes each correct response
carries equal evidential weight. \textcite{koedinger2007} demonstrate that this
assumption fails when scaffolding is available: correct responses following
worked examples are substantially less informative about latent knowledge than
unaided correct responses. This project modifies BKT by scaling the effective
learning gain by a credit factor $\text{cf} \in \{1.0, 0.9, 0.7, 0.5\}$
depending on the maximum hint level revealed:

\begin{equation}
  p(L_{n+1})^* = p(L_n \mid \text{obs}) + \bigl(1 - p(L_n \mid \text{obs})\bigr) \cdot p(T) \cdot \text{cf}
  \label{eq:bkt_credit}
\end{equation}

The credit factor values follow the empirical estimates reported by
\textcite{koedinger2007} from their reanalysis of three Cognitive Tutor datasets.
The asymmetry (0.7 for procedural hints vs. 0.5 for worked examples) reflects
the stepped scaffolding hierarchy: a worked example fully discloses the solution,
so correctness carries minimal diagnostic signal.

\subsection{Deep Knowledge Tracing and Neural Approaches}
\label{subsec:lr_dkt}

\textcite{piech2015} introduced Deep Knowledge Tracing (DKT), replacing BKT's
explicit probabilistic model with a Long Short-Term Memory (LSTM) recurrent
neural network. Trained on a sequence of (KC, correctness) pairs, DKT
outperforms BKT on held-out prediction tasks across several benchmark datasets,
achieving Area Under the Curve (AUC) improvements of approximately 6\% over BKT.

However, \textcite{khajah2016} demonstrate that a substantial portion of DKT's
predictive advantage over BKT disappears when BKT is extended with temporal
features and per-student priors: the performance gap reduces from 6\% to 1--2\%.
More critically for this project, DKT sacrifices interpretability entirely: the
hidden state vector of an LSTM has no correspondence to psychologically
meaningful constructs, and adaptive decisions cannot be traced to specific
knowledge state estimates. \textcite{matos2025} confirm that in small-scale
educational deployments (the context of this project), rule-based and
probabilistic systems remain more appropriate than deep learning approaches
precisely because their transparency supports both learner trust and academic
scrutiny. DKT is therefore unsuitable for a dissertation project requiring
full accountability of every adaptive decision.

\subsection{Hybrid IRT + BKT: Rationale and Prior Work}
\label{subsec:lr_hybrid}

The hybrid approach integrates IRT's item-level psychometric precision with
BKT's temporal, per-KC knowledge state modelling. Each algorithm addresses the
other's primary limitation: BKT provides no item-level discrimination, which
causes all questions of equal difficulty to be treated identically regardless of
their discriminating power; IRT provides no per-KC mastery tracking, which
prevents mastery-gated progression and prerequisite enforcement.

\textcite{kaser2017} confirm across 1.2 million student-problem interactions
that the predictive advantage of IRT over Elo is modest (82\% vs. 78\%
prediction accuracy), but that the critical value of the hybrid architecture
lies not in marginal accuracy improvement but in the simultaneous modelling of
item properties and temporal knowledge dynamics.

\textcite{desmarais2012} review 40 learner modelling systems and conclude that
hybrid probabilistic approaches consistently outperform single-algorithm systems
in structured multi-KC curricula. The algorithm selection applied to this
project is summarised in Table~\ref{tab:algo_comparison}.

\begin{table}[H]
\centering
\caption{Comparative evaluation of adaptive algorithms against project requirements.}
\label{tab:algo_comparison}
\renewcommand{\arraystretch}{1.3}
\begin{tabularx}{\textwidth}{lXXXX}
\toprule
\textbf{Criterion} & \textbf{Elo} & \textbf{IRT 3PL} & \textbf{BKT} & \textbf{IRT+BKT Hybrid} \\
\midrule
Data requirements & Low & High & Medium & Medium (with priors) \\
Per-KC mastery tracking & No & No & Yes & Yes \\
Prerequisite enforcement & No & No & Yes & Yes \\
Item discrimination modelling & No & Yes & No & Yes \\
Knowledge decay detection & No & No & Partial & Yes (via decay flag) \\
Interpretability & High & High & High & High \\
Calibration stability (small $n$) & High & Low & Medium & Medium \\
\bottomrule
\end{tabularx}
\end{table}

% ─────────────────────────────────────────────────────────────────────────────
\section{Computerised Adaptive Testing}
\label{sec:lr_cat}
% ─────────────────────────────────────────────────────────────────────────────

\subsection{Principles and Item Selection}
\label{subsec:lr_cat_principles}

Computerised Adaptive Testing (CAT) is the application of IRT to sequential item
selection, in which each successive item is chosen based on all previous
responses rather than from a fixed predetermined sequence \parencite{wainer2000}.
The foundational principle is maximum information selection: the optimal item at
any point in a test is the item maximising Fisher information at the learner's
current ability estimate. Fisher information for the 3PL model is:

\begin{equation}
  I(\theta) = \frac{a^2 \cdot P(\theta)(1-P(\theta))}{\left(\dfrac{P(\theta)-c}{1-c}\right)^2}
  \label{eq:fisher_information}
\end{equation}

\noindent where $P(\theta)$ is the item characteristic function evaluated at the
current $\theta$ estimate. Information is maximised at the point on the ability
scale where the item provides maximum discriminating power, which for 3PL items
occurs slightly below $b$ (the difficulty parameter) due to the non-zero lower
asymptote $c$. Maximum information selection is the standard CAT item selection
criterion \parencite{weiss1982} and is implemented directly in this system's
question selection algorithm.

\subsection{Ability Estimation: EAP vs.\ MLE}
\label{subsec:lr_eap}

Two primary methods exist for estimating $\theta$ from a response sequence:
Maximum Likelihood Estimation (MLE) and Expected A Posteriori (EAP) estimation.
MLE finds the $\hat{\theta}$ maximising the log-likelihood:

\begin{equation}
  \hat{\theta}_{\text{MLE}} = \arg\max_\theta \sum_{i=1}^{n} \left[ x_i \ln P_i(\theta) + (1-x_i) \ln (1 - P_i(\theta)) \right]
  \label{eq:mle}
\end{equation}

MLE is asymptotically unbiased but has a critical limitation: it is undefined
when a learner answers all items correctly or all items incorrectly, which is
common in short sessions. EAP estimation integrates the posterior distribution
over a discrete grid of ability values:

\begin{equation}
  \hat{\theta}_{\text{EAP}} = \frac{\displaystyle\sum_k \theta_k \cdot L(\theta_k) \cdot \pi(\theta_k)}{\displaystyle\sum_k L(\theta_k) \cdot \pi(\theta_k)}
  \label{eq:eap}
\end{equation}

\noindent where $L(\theta_k)$ is the likelihood of the observed response
pattern at grid point $\theta_k$ and $\pi(\theta_k)$ is the prior probability.
EAP always produces a finite estimate and incorporates prior uncertainty through
the prior distribution $\pi(\theta)$, making it more robust for short sessions
of 20--27 items \parencite{bock1982}.

This project implements EAP over a grid $\theta \in [-4, 4]$ with step $0.1$
(81 grid points), using a standard normal prior $\theta_0 \sim \mathcal{N}(-0.780, 0.543)$,
where the initial mean of $-0.780$ corresponds to the 20th percentile of the
normalised ability distribution, reflecting the conservative assumption that a
new learner is likely below average relative to the general population.
\textcite{baker2004} confirm that EAP with a well-specified prior outperforms
MLE in short-test conditions on both bias and mean squared error criteria.

% ─────────────────────────────────────────────────────────────────────────────
\section{Empirical Effectiveness of Intelligent Tutoring Systems}
\label{sec:lr_effectiveness}
% ─────────────────────────────────────────────────────────────────────────────

\subsection{Meta-analyses and Effect Sizes}
\label{subsec:lr_meta}

\textcite{vanlehn2011} provides the most comprehensive evaluation of ITS
effectiveness, synthesising 60 controlled studies and reporting the following
effect size hierarchy:

\begin{table}[H]
\centering
\caption{Effect size comparison across tutoring modalities \parencite{vanlehn2011}.}
\label{tab:effect_sizes}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{lc}
\toprule
\textbf{Tutoring modality} & \textbf{Effect size ($d$)} \\
\midrule
Human one-to-one tutoring & 0.79 \\
Step-based ITS & 0.76 \\
Answer-based tutoring systems & 0.31 \\
No-tutoring control & 0.00 (baseline) \\
\bottomrule
\end{tabular}
\end{table}

The near-equivalence of step-based ITS and human tutoring ($d = 0.76$ vs.
$d = 0.79$) is the empirical foundation justifying the development of this
system. \textcite{kulik2016} corroborate these findings across 50 experimental
studies, reporting a weighted average effect size of $d = 0.42$ for adaptive
systems over conventional instruction, with systems implementing continuous
real-time adaptation ($d = 0.66$) substantially outperforming discrete checkpoint
systems ($d = 0.28$). VanLehn's benchmark of $d = 0.76$ is adopted as the
primary comparison benchmark for this project's evaluation.

\subsection{Effectiveness in Non-STEM Domains}
\label{subsec:lr_nonstems}

The preponderance of ITS research addresses procedural knowledge domains,
particularly mathematics and programming, where objective performance metrics are
readily available and knowledge component boundaries are well-defined. Evidence
for ITS effectiveness in declarative-knowledge domains remains comparatively
sparse. \textcite{steenbergen2013} confirm strong ITS effects on mathematical
learning ($d = 0.36$ across 26 studies) but note that ``the generalisability of
these findings to non-mathematical domains remains an open empirical question''
(p. 984).

Geography education presents an intermediate epistemological structure. Factual
knowledge (capital cities, river systems, physical landscape features) admits
objective assessment. Procedural knowledge (interpreting climate graphs,
analysing population data) requires application of geographic frameworks.
Conceptual understanding (explaining spatial inequality, evaluating sustainability
arguments) involves higher-order reasoning. \textcite{lambert2010} characterise
Geography as a ``thinking subject'', emphasising that effective Geography
education requires learners to develop not merely factual recall but geographical
imaginations: the capacity to interpret spatial patterns and processes.
This epistemological complexity makes Geography a meaningful and underexplored
testbed for ITS research.

\textcite{bednarz2013} identify significant underrepresentation of UK Geography
specifically in adaptive learning research. The predominance of North American
and East Asian studies \parencite{matos2025} means that the particular pedagogical
context of UK KS3/GCSE Geography, including the AQA/Edexcel specification
structure, the emphasis on fieldwork and geographical enquiry, and the UK
curriculum's distinctive treatment of human-physical geography integration,
remains insufficiently studied.

\subsection{Long-Term Retention}
\label{subsec:lr_retention}

A persistent methodological limitation in ITS research is the near-exclusive
reliance on immediate post-test performance as the outcome measure. Most studies
measure learning gains within a single session or immediately following
intervention, providing no evidence about knowledge retention over ecologically
relevant timescales. \textcite{cepeda2006} demonstrate that the relative
effectiveness of learning strategies changes substantially when measured at
delayed retention intervals: massed practice often outperforms spaced practice
on immediate tests but is consistently outperformed by spaced practice at delays
of one week or more.

The inclusion of a 7-day delayed retention assessment in this project's
evaluation protocol directly addresses this methodological gap, providing
evidence about whether the adaptive system's learning gains are retained over
a period consistent with educational relevance, rather than merely reflecting
short-term performance improvement.

% ─────────────────────────────────────────────────────────────────────────────
\section{Implementation Challenges}
\label{sec:lr_challenges}
% ─────────────────────────────────────────────────────────────────────────────

\subsection{The Cold-Start Problem}
\label{subsec:lr_coldstart}

The cold-start problem describes the challenge of personalising effectively
before sufficient learner data has accumulated. A new learner who has not
interacted with the system provides no response data from which to estimate
ability, making the first several interactions necessarily suboptimal.
\textcite{matos2025} identify the cold-start problem as a ``significant
implementation gap'' in real-world adaptive system deployments.

This project addresses the cold-start problem through pre-test seeding: before
the adaptive session begins, a 13-item diagnostic assessment spanning all KCs
at Bloom Level 1 provides initial ability evidence. The resulting response
pattern is converted to a Laplace-smoothed logit estimate:

\begin{equation}
  \theta_{\text{seed}} = \frac{\ln\!\left(\dfrac{n_{\text{correct}} + 0.5}{n_{\text{incorrect}} + 0.5}\right)}{1.7}
  \label{eq:theta_seed}
\end{equation}

\noindent where the 1.7 scaling factor is the standard IRT logistic approximation
mapping log-odds to the normal-ogive $D$ parameter \parencite{lord1980}. Laplace
smoothing prevents boundary estimates at perfect or zero scores. The seed
estimate is bounded at $\theta \in [-2.5, 2.5]$ and used to set the session
starting theta, with a reduced prior standard deviation of 0.35 (vs.\ the default
0.543) reflecting the additional information from the pre-test.

\subsection{IRT Calibration with Small Samples}
\label{subsec:lr_calibration}

Traditional IRT calibration via expectation-maximisation or marginal maximum
likelihood requires response matrices of 500+ observations per item for stable
parameter recovery, particularly for the 3PL model where the $c$ parameter is
notoriously difficult to estimate with sparse data \parencite{lord1980}. This
project's pilot study involves a small number of participants, making full 3PL
calibration with freely estimated parameters infeasible.

The calibration strategy adopted is a constrained two-stage pipeline: item
parameters are estimated using the \texttt{mirt} package in R \parencite{chalmers2012}
with the guessing parameter $c$ fixed at 0.25 (its theoretical value for
four-option items), and the calibration is verified against pilot data using
convergence diagnostics. If 3PL estimation proves unstable (RMSEA $> 0.10$ or
non-convergence within 500 EM cycles), a fallback to the Rasch (1PL) model is
pre-specified. This decision is documented explicitly in the project's risk
register and constitutes a defensible adaptive methodology, not a failure: the
Rasch model's parsimony is a genuine advantage in small-sample conditions
\parencite{embretson2000}.

\subsection{Usability and User Experience Design}
\label{subsec:lr_usability}

\textcite{bangor2008} established the System Usability Scale (SUS) as the
standard metric for evaluating educational technology usability. Products scoring
above 68 are considered above average; scores above 85 indicate excellent
usability. Their analysis of 206 studies found adaptive systems typically score
15--20 points lower than conventional interfaces due to increased navigational
complexity and unpredictable content sequencing.

\textcite{mayer2009} provides the theoretical foundation for minimising extraneous
cognitive load through multimedia design principles: coherence (remove extraneous
material), signalling (highlight the organisation), redundancy (avoid
representing the same information in multiple formats simultaneously), and
segmentation (allow learner-controlled pacing). These principles directly inform
the interface design choices in this project: the five-tab session summary
provides coherence by separating analytical information into distinct categories;
the first-visit tour provides signalling by explaining the adaptive engine's
decisions; and the pause-and-resume mechanism implements segmentation by allowing
learner-controlled session pacing.

% ─────────────────────────────────────────────────────────────────────────────
\section{Ethical Considerations}
\label{sec:lr_ethics}
% ─────────────────────────────────────────────────────────────────────────────

\subsection{Algorithmic Transparency and Accountability}
\label{subsec:lr_transparency}

\textcite{baker2021} identify three primary ethical dimensions of adaptive
educational systems. First, \textit{transparency}: learners and educators have a
right to understand why the system presents particular content. Black-box
adaptation, in which the learner experiences a changing sequence of questions
with no explanation of the underlying logic, erodes trust and agency.
\textcite{murray1999} argues that for educational systems to be accepted by
learners, the system's reasoning must be at least partially legible.

This project addresses transparency through the session summary dashboard's
five-tab interface: the Trajectory tab displays the learner's theta progression
across a session, the Skills tab shows per-KC BKT mastery probabilities, and
the Misconceptions tab presents the most frequent wrong-answer attractors per
KC. These visualisations make the adaptive engine's internal state visible and
interpretable, enabling learners to understand why the system selected specific
questions and to identify their own knowledge gaps.

\subsection{Data Protection and GDPR Compliance}
\label{subsec:lr_gdpr}

\textcite{voigt2017} provide a comprehensive treatment of the General Data
Protection Regulation's (GDPR) implications for educational technology. The
critical GDPR principles relevant to this project are: \textit{data minimisation}
(collect only the personal data necessary for the stated purpose);
\textit{purpose limitation} (use collected data only for the purpose for which
consent was obtained); and \textit{storage limitation} (retain personal data
only for as long as necessary).

This system implements GDPR compliance through: server-side validated consent
flows presented before any data collection begins; storage of only
session-level performance data (no geolocation, no device fingerprinting, no
persistent tracking beyond the study period); JWT-based authentication with
bcrypt password hashing following the authentication security guidelines of
\textcite{nist2017}; and rate limiting on authentication endpoints to prevent
credential stuffing attacks.

\subsection{Algorithmic Bias and Equity}
\label{subsec:lr_bias}

\textcite{baker2021} document that adaptive systems can inadvertently reinforce
educational inequities through differential error rates across demographic groups.
In Geography specifically, prior access to geographical experiences (international
travel, diverse urban environments) may correlate with item performance in ways
unrelated to curricular knowledge, producing systematically biased ability
estimates for learners from less privileged backgrounds.

This limitation is acknowledged in this project's evaluation protocol as a
constraint on the generalisability of findings: the participant pool (University
of Hull undergraduate and postgraduate students) is not representative of the
KS3/GCSE student population the system targets. The evaluation findings should
therefore be interpreted as proof-of-concept validation in a convenience sample
rather than as evidence of universal effectiveness. Bias auditing procedures
are identified as a priority for any production deployment of the system.

% ─────────────────────────────────────────────────────────────────────────────
\section{Research Gaps and Project Contribution}
\label{sec:lr_gaps}
% ─────────────────────────────────────────────────────────────────────────────

The literature reviewed above reveals five significant gaps that this project
directly addresses.

\textbf{Gap 1: ITS application to UK Geography.}
No published ITS system targets UK KS3/GCSE Geography specifically. The
overwhelming majority of adaptive learning research addresses mathematics,
programming, and North American science curricula. This project provides a
proof-of-concept adaptive system for the UK Geography curriculum, extending the
empirical evidence base for ITS effectiveness beyond traditional STEM domains.

\textbf{Gap 2: Hybrid IRT+BKT implementation with open source accessibility.}
While hybrid approaches are theoretically advocated \parencite{desmarais2012,
matos2025}, few open-source implementations exist that combine IRT's item-level
precision with BKT's per-KC mastery tracking in a production-deployed web
application. This project provides a publicly accessible implementation at
\url{https://github.com/dlalloyd/E-Learning-Final-Project}.

\textbf{Gap 3: Bloom-gated adaptive progression.}
Existing ITS systems typically sequence items by IRT difficulty alone, without
enforcing cognitive-level prerequisites. This project implements Bloom-gated
selection: the adaptive engine enforces that learners consolidate recall before
encountering analysis-level items, implementing the taxonomy's hierarchical
structure as a computational constraint.

\textbf{Gap 4: Long-term retention evaluation.}
The near-universal reliance on immediate post-test outcomes in ITS evaluation
research is a methodological limitation noted by multiple reviewers
\parencite{cepeda2006, vanlehn2011}. This project's 7-day delayed retention
assessment provides evidence on the persistence of adaptive learning gains.

\textbf{Gap 5: Interpretability under real-time adaptation.}
The interpretability-efficacy trade-off identified by \textcite{matos2025},
in which more powerful adaptive algorithms tend to be less interpretable,
is addressed through the combination of an interpretable probabilistic model
(IRT+BKT, where every adaptive decision is traceable to specific parameter values)
and a transparent visualisation layer (the five-tab session summary dashboard).

% ─────────────────────────────────────────────────────────────────────────────
\section{Future Directions: Cross-Domain Knowledge Transfer}
\label{sec:lr_future}
% ─────────────────────────────────────────────────────────────────────────────

The system is architecturally domain-agnostic: the adaptive engine operates on
any content that can be expressed as Knowledge Components with prerequisite
edges, IRT-parameterised question variants, and BKT mastery states. Geography
was chosen as the validation domain for this project because its declarative-
knowledge structure admits objective assessment and because the domain is
underrepresented in existing ITS research. However, the architecture is designed
to support extension to additional domains without modification of the core
adaptive engine.

The most theoretically significant extension is cross-domain knowledge transfer
\parencite{singley1989}: the identification of shared cognitive skills between
domains and the adaptive allocation of practice to transfer-relevant knowledge
components. \textcite{koedinger2012} demonstrate through the
Knowledge-Learning-Instruction (KLI) framework that learning outcomes depend
not merely on domain knowledge but on the cognitive features of instructional
events: practice that develops both a declarative knowledge fact and the
procedural skill of applying it in novel contexts produces superior transfer.

In the context of this system's architecture, cross-domain transfer would be
implemented through: shared KC representations across domain boundaries (for
example, the spatial reasoning required to interpret population distribution maps
in Geography shares cognitive structure with statistical reasoning in Mathematics);
inter-domain prerequisite edges encoding these structural relationships; and
cross-domain BKT parameters allowing learning in one domain to inform the prior
estimates in another. \textcite{brusilovsky2004} identify adaptive sequencing
across domain boundaries as an open research challenge; this project's
domain-agnostic architecture provides the technical foundation for future
investigation of this problem.

% ─────────────────────────────────────────────────────────────────────────────
\section{Conclusions}
\label{sec:lr_conclusion}
% ─────────────────────────────────────────────────────────────────────────────

The literature establishes that adaptive intelligent tutoring systems can
approach human tutoring effectiveness when designed with appropriate interaction
granularity, principled algorithm selection, and transparent feedback mechanisms.
The hybrid IRT+BKT approach is theoretically optimal for structured multi-KC
curricula with prerequisite dependencies, and the UK Geography domain represents
a meaningful and underexplored validation context.

The following design principles, derived from this review, inform the system
architecture presented in Chapter~\ref{ch:methodology}:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{ZPD operationalisation}: target $P(\theta) \approx 0.70$ for
  item selection, maintaining challenge within the zone of proximal development.
  \item \textbf{Mastery gating}: require $p(\text{Learned}) \geq 0.80$ per KC
  before permitting progression to dependent KCs.
  \item \textbf{Bloom-cognitive scaffolding}: enforce Bloom-level prerequisites
  as selection gates, not merely as difficulty labels.
  \item \textbf{EAP estimation}: use Expected A Posteriori ability estimation
  over a discretised grid to ensure stability in short sessions.
  \item \textbf{Credit-adjusted BKT}: reduce the evidential weight of hint-assisted
  correct responses using graduated credit factors.
  \item \textbf{Successive relearning}: prioritise re-presentation of decayed KCs
  through discrimination-boosted selection.
  \item \textbf{Transparent adaptation}: surface internal model states to the
  learner through visualisation, not concealment.
\end{enumerate}

These principles collectively instantiate an adaptive system that is
simultaneously pedagogically grounded, psychometrically rigorous, interpretable
to learners and examiners, and architecturally extensible to future domains.
