\chapter{Methodology}
\label{ch:methodology}

% ─────────────────────────────────────────────────────────────────────────────
\section{Research Design}
\label{sec:meth_design}
% ─────────────────────────────────────────────────────────────────────────────

This project adopts a \textit{Design Science Research} (DSR) methodology
\parencite{hevner2004}, which prescribes the iterative construction and
evaluation of an artefact intended to solve a practical problem. DSR is
appropriate when the research goal is not merely to observe an existing
phenomenon but to engineer a system that demonstrably improves upon the
current state of practice. The artefact produced is GeoMentor: a
production-deployed adaptive intelligent tutoring system for UK KS3/GCSE
Geography implementing a hybrid IRT 3PL + BKT algorithm.

The evaluation component follows a \textbf{within-subjects quasi-experimental
design}. Each participant experiences both a baseline linear condition and the
adaptive condition within a single extended session, eliminating individual
ability as a confound. The absence of randomised assignment to groups (due to
the convenience sample and within-subjects structure) classifies the design as
quasi-experimental rather than a true randomised controlled trial. This is a
documented limitation acknowledged in Section~\ref{sec:meth_evaluation}.

The scope was deliberately concentrated on a single Knowledge Component domain
(UK physical geography and climate, 13 KCs covering spatial recall, climate
processes, and integrative application within the KS3/GCSE Geography
specification) rather than the multi-domain ambition of the initial proposal. This decision, taken following technical risk assessment in
the second project sprint (February 2026), strengthens rather than weakens the
evaluation claims: validation depth within a single well-specified domain
produces more defensible psychometric evidence than shallow validation across
multiple domains. Cross-Domain Knowledge Transfer remains the primary
architectural stretch goal and is addressed in the system's domain-agnostic
design (Section~\ref{sec:meth_kc}) and discussed as future work in
Chapter~\ref{ch:discussion}.

% ─────────────────────────────────────────────────────────────────────────────
\section{Knowledge Representation}
\label{sec:meth_kc}
% ─────────────────────────────────────────────────────────────────────────────

\subsection{Knowledge Component Ontology}
\label{subsec:meth_kc_ontology}

The curriculum is decomposed into 13 Knowledge Components (KCs) covering UK
physical geography and climate. A KC is defined as a unit of knowledge or
skill that can be independently assessed and for which mastery can be
operationally defined \parencite{koedinger2012}. The domain was scoped to UK
climate systems after a Sprint~5 risk assessment identified that calibrating
IRT parameters across the full AQA specification breadth was infeasible
within the available development timeline; this decision is further justified
in Section~\ref{sec:meth_evaluation}. Table~\ref{tab:kc_table} lists all
13 KCs with their Bloom cognitive level ceiling and BKT prior parameters
as implemented in \texttt{DEFAULT\_BKT\_PARAMS} in \texttt{src/lib/algorithms/bkt.ts}.

\begin{table}[H]
\centering
\caption{Knowledge Component ontology: 13 KCs with BKT prior parameters.
  $p(S)$ and $p(G)$ are KC-specific; $p(G) = 0.25$ for all KCs (fixed,
  four-option MCQ). KC identifiers match the live \texttt{DEFAULT\_BKT\_PARAMS}.}
\label{tab:kc_table}
\renewcommand{\arraystretch}{1.25}
\begin{tabularx}{\textwidth}{lXcccc}
\toprule
\textbf{\#} & \textbf{Knowledge Component} & \textbf{Bloom} &
  $p(L_0)$ & $p(T)$ & $p(S)$ \\
\midrule
\multicolumn{6}{l}{\textit{Level 1: Remembering}} \\
1  & UK Capitals (\texttt{UK\_capitals})                   & L1 & 0.60 & 0.25 & 0.08 \\
2  & County Locations (\texttt{UK\_county\_locations})     & L1 & 0.45 & 0.20 & 0.12 \\
3  & UK Rivers (\texttt{UK\_rivers})                       & L1 & 0.55 & 0.22 & 0.10 \\
4  & UK Mountains (\texttt{UK\_mountains})                 & L1 & 0.50 & 0.18 & 0.10 \\
5  & National Parks (\texttt{UK\_national\_parks})         & L1 & 0.45 & 0.16 & 0.12 \\
\multicolumn{6}{l}{\textit{Level 2: Understanding}} \\
6  & Westerly Winds \& Rainfall (\texttt{westerly\_winds\_rainfall})    & L2 & 0.25 & 0.18 & 0.12 \\
7  & Maritime vs Continental (\texttt{maritime\_continental})           & L2 & 0.20 & 0.18 & 0.12 \\
8  & Pennines Rain Shadow (\texttt{pennines\_rain\_shadow})             & L2 & 0.20 & 0.15 & 0.15 \\
9  & North Atlantic Drift (\texttt{north\_atlantic\_drift})             & L2 & 0.22 & 0.18 & 0.12 \\
10 & Continental Effect (\texttt{continental\_effect})                  & L2 & 0.18 & 0.15 & 0.15 \\
\multicolumn{6}{l}{\textit{Level 3: Applying}} \\
11 & Climate Classification (\texttt{climate\_classification})          & L3 & 0.15 & 0.15 & 0.18 \\
12 & Climate Change Application (\texttt{climate\_change\_application}) & L3 & 0.12 & 0.15 & 0.18 \\
13 & Flood Risk Integration (\texttt{flood\_risk\_integration})         & L3 & 0.10 & 0.12 & 0.20 \\
\bottomrule
\end{tabularx}
\end{table}

Prior probabilities $p(L_0)$ are highest for Level~1 KCs (0.45--0.60),
reflecting realistic recall knowledge for adult learners, and lowest for
Level~3 KCs (0.10--0.15), reflecting the synthesis demands of integrative
items. Within Level~1, UK~Capitals receives the highest $p(L_0) = 0.60$
(only four facts to recall) while County~Locations and National~Parks receive
lower priors (0.45), reflecting the larger recall set of 47 counties and
15 parks respectively. The learning transition $p(T)$ follows the same
gradient: $p(T) = 0.25$ for UK~Capitals (rapid mastery expected),
$p(T) = 0.16$ for National~Parks (slow acquisition over a large item set),
and $p(T) = 0.12$ for Flood~Risk~Integration (synthesis-level knowledge
acquired incrementally). The slip parameter $p(S)$ scales with Bloom level
(0.08--0.12 at L1, 0.12--0.15 at L2, 0.18--0.20 at L3), reflecting that
higher-order items carry greater response uncertainty even for knowledgeable
learners \parencite{pardos2011}. The guess parameter $p(G) = 0.25$ is fixed
for all KCs, corresponding to the uniform four-option MCQ format.

\subsection{Prerequisite Graph}
\label{subsec:meth_prerequisite}

KCs are not independent; mastery of foundational KCs is necessary for
productive engagement with dependent KCs. The prerequisite graph encodes these
structural dependencies and governs mastery-gated KC progression. The graph
is implemented as a directed acyclic graph (DAG) with 12 edges seeded into
the \texttt{PrerequisiteEdge} table at deployment time and queried by the
adaptive engine at each question selection step.

Five Level~1$\to$Level~2 edges gate the climate-process KCs behind the
spatial KCs that contextualise them (for example, understanding westerly
winds presupposes knowing where the relevant counties and coastlines are).
One Level~2$\to$Level~2 edge encodes the causal dependency of the Pennines
Rain~Shadow on westerly wind patterns: a learner cannot meaningfully reason
about orographic enhancement without first understanding the prevailing air
mass trajectory. Five Level~2$\to$Level~3 edges and one direct
Level~1$\to$Level~3 edge gate the integrative KCs.

All mastery gates use a uniform threshold of $p(\text{Learned}) \geq 0.80$,
consistent with the BKT mastery convention established by
\textcite{corbett1995} and used in empirically validated ITS deployments
\parencite{vanlehn2011}. UK~Capitals and National~Parks have no outgoing
edges: they are assessed throughout sessions but do not gate any higher-order
KCs.

\begin{figure}[H]
\centering
\begin{tikzpicture}[
  kc/.style={rectangle, rounded corners=4pt, draw=black!70,
    text width=2.5cm, align=center, font=\scriptsize,
    minimum height=0.9cm, inner sep=3pt},
  l1/.style={kc, fill=green!7, draw=green!55!black},
  l2/.style={kc, fill=blue!7, draw=blue!45},
  l3/.style={kc, fill=orange!8, draw=orange!65!black},
  arr/.style={-{Latex[length=2mm]}, gray!65, thick}
]

%% Level 1: Remembering (y = 4.5) -- five spatial/locational KCs
\node[l1] (caps)    at ( 0.0, 4.5) {UK\\Capitals};
\node[l1] (county)  at ( 2.5, 4.5) {County\\Locations};
\node[l1] (rivers)  at ( 5.0, 4.5) {UK\\Rivers};
\node[l1] (mnts)    at ( 7.5, 4.5) {UK\\Mountains};
\node[l1] (parks)   at (10.0, 4.5) {National\\Parks};

%% Level 2: Understanding (y = 2.5) -- five climate-process KCs
\node[l2] (westerly)    at ( 0.0, 2.5) {Westerly\\Winds};
\node[l2] (maritime)    at ( 2.5, 2.5) {Maritime vs\\Continental};
\node[l2] (pennines)    at ( 5.0, 2.5) {Pennines Rain\\Shadow};
\node[l2] (natldrift)   at ( 7.5, 2.5) {N.\ Atlantic\\Drift};
\node[l2] (continental) at (10.0, 2.5) {Continental\\Effect};

%% Level 3: Applying (y = 0.5) -- three integrative KCs
\node[l3] (climclass)  at ( 2.5, 0.5) {Climate\\Classification};
\node[l3] (climchange) at ( 6.5, 0.5) {Climate Change\\Application};
\node[l3] (flood)      at (10.0, 0.5) {Flood Risk\\Integration};

%% L1 → L2 (five edges)
\draw[arr] (county) to[out=200,in=80]  (westerly);
\draw[arr] (county) --                 (maritime);
\draw[arr] (county) to[out=350,in=90]  (continental);
\draw[arr] (mnts)   --                 (pennines);
\draw[arr] (rivers) --                 (natldrift);

%% L2 → L2 (one edge: westerly winds enable rain shadow understanding)
\draw[arr] (westerly) -- (pennines);

%% L2 → L3 (four edges)
\draw[arr] (maritime)    -- (climclass);
\draw[arr] (continental) to[out=215,in=5]   (climclass);
\draw[arr] (natldrift)   -- (climchange);
\draw[arr] (pennines)    to[out=310,in=150]  (flood);

%% L1 → L3 (one cross-level edge)
\draw[arr] (rivers) to[out=265,in=130] (flood);

%% L2 → L3 (westerly winds → flood risk via rainfall loading)
\draw[arr] (westerly) to[out=320,in=185] (flood);

\end{tikzpicture}
\caption{KC prerequisite DAG: 13 nodes across three Bloom levels
  (green~=~L1 Remembering, blue~=~L2 Understanding, orange~=~L3 Applying).
  Arrows indicate structural dependency (prerequisite $\to$ dependent);
  mastery gate $p(\text{Learned}) \geq 0.80$ on the prerequisite KC before
  dependent-KC items unlock. KCs without outgoing edges (UK~Capitals,
  National~Parks) are assessed throughout but do not gate higher-order KCs.
  12 edges total; 5 L1$\to$L2, 1 L2$\to$L2, 5 L2$\to$L3, 1 L1$\to$L3.}
\label{fig:kc_dag}
\end{figure}

% ─────────────────────────────────────────────────────────────────────────────
\section{Adaptive Algorithm Design}
\label{sec:meth_algorithms}
% ─────────────────────────────────────────────────────────────────────────────

\subsection{IRT 3PL Model}
\label{subsec:meth_irt}

The probability of a correct response from a learner with latent ability
$\theta$ on item $i$ is modelled by the three-parameter logistic (3PL)
function \parencite{birnbaum1968}:

\begin{equation}
  P(X=1 \mid \theta) = c_i + \frac{1-c_i}{1 + e^{-a_i(\theta - b_i)}}
  \label{eq:irt_3pl_meth}
\end{equation}

where $a_i \in [0.5, 3.0]$ is the discrimination parameter (steepness of the
Item Characteristic Curve), $b_i \in [-3, 3]$ is the difficulty parameter
(logit scale), and $c_i$ is the pseudo-guessing lower asymptote (fixed at
$0.25$ for all four-option MCQs). The 3PL model is calibrated offline using
the \texttt{mirt} package in R \parencite{chalmers2012} and the resulting
parameter estimates are serialised to a JSON configuration file consumed by
the Node.js runtime (see Section~\ref{subsec:meth_calibration}).

\subsection{EAP Ability Estimation}
\label{subsec:meth_eap}

Learner ability $\theta$ is updated after each response using Expected A
Posteriori (EAP) estimation, preferred over MLE for short sessions due to
its stability when response patterns are monotone (see
Section~\ref{subsec:lr_eap}). EAP integrates over a discrete grid:

\begin{equation}
  \hat{\theta}_{\text{EAP}} = \frac{\displaystyle\sum_{k=1}^{K}
    \theta_k \cdot L(\theta_k) \cdot \pi(\theta_k)}
    {\displaystyle\sum_{k=1}^{K} L(\theta_k) \cdot \pi(\theta_k)}
  \label{eq:eap_meth}
\end{equation}

where $K = 81$ grid points span $[-4, 4]$ with step $0.1$. The prior
$\pi(\theta_k) = \mathcal{N}(\mu_0, \sigma_0)$ is initialised at
$\mu_0 = -0.780$ (20th percentile, reflecting conservative prior for a new
learner) and $\sigma_0 = 0.543$. Following pre-test seeding
(Section~\ref{sec:meth_pretest}), the prior is updated to incorporate the
pre-test evidence with a tightened standard deviation of $0.35$.

\subsection{Bayesian Knowledge Tracing}
\label{subsec:meth_bkt}

Per-KC knowledge state is maintained as a BKT posterior \parencite{corbett1995}.
After each observed response, the posterior is updated via Bayes' theorem:

\begin{equation}
  p(L_n \mid \text{obs}_n) = \frac{P(\text{obs}_n \mid L_n) \cdot p(L_n)}
    {P(\text{obs}_n \mid L_n) \cdot p(L_n) + P(\text{obs}_n \mid \neg L_n)
    \cdot (1 - p(L_n))}
  \label{eq:bkt_update_meth}
\end{equation}

where $P(\text{correct} \mid L_n) = 1 - p(S)$ and
$P(\text{correct} \mid \neg L_n) = p(G)$. The learning transition then
produces the updated prior for the next observation:

\begin{equation}
  p(L_{n+1}) = p(L_n \mid \text{obs}_n) + \bigl(1 - p(L_n \mid \text{obs}_n)\bigr)
    \cdot p(T)
  \label{eq:bkt_transition_meth}
\end{equation}

\subsubsection*{Credit Factor Modification}

Standard BKT treats all correct responses as equally informative. This
project modifies the learning transition by a credit factor $\text{cf}$
reflecting the scaffolding level at which the correct response was obtained:

\begin{equation}
  p(L_{n+1})^* = p(L_n \mid \text{obs}_n) + \bigl(1 - p(L_n \mid
    \text{obs}_n)\bigr) \cdot p(T) \cdot \text{cf}
  \label{eq:bkt_credit_meth}
\end{equation}

\begin{table}[H]
\centering
\caption{Credit factor values by maximum hint level accessed per question.}
\label{tab:credit_factors}
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{clc}
\toprule
\textbf{Max hint level} & \textbf{Scaffolding type} & \textbf{cf} \\
\midrule
0 & No hint accessed (unaided)         & 1.00 \\
1 & Conceptual reframe                 & 0.90 \\
2 & Procedural strategy hint           & 0.70 \\
3 & Worked example (full solution)     & 0.50 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Question Selection Algorithm}
\label{subsec:meth_selection}

At each step the adaptive engine selects the next question from the eligible
item pool via the following procedure (Figure~\ref{fig:selection_flowchart}):

\begin{enumerate}[leftmargin=2em]
  \item \textbf{Mastery gate}: exclude all items whose KC has not yet been
        unlocked (i.e., a prerequisite KC has $p(\text{Learned}) < 0.80$).
  \item \textbf{Bloom gate}: exclude items whose Bloom level exceeds the
        learner's current escalation threshold for that KC:
        \begin{itemize}
          \item L2 unlocks when $p(\text{Learned}) \geq 0.40$ \textit{or} a
                direct prerequisite KC exceeds $0.50$
          \item L3 unlocks when $p(\text{Learned}) \geq 0.60$
        \end{itemize}
  \item \textbf{Deduplication}: exclude items whose variant has been seen by
        this learner in any prior session (\texttt{UserVariantSeen} table).
  \item \textbf{Decay boost}: for KCs flagged as decayed (cross-session
        $p(\text{Learned})$ decline), multiply the Fisher information by $1.5$.
  \item \textbf{ZPD targeting}: from the remaining eligible pool, select the
        item maximising Fisher information at the current $\hat{\theta}$:
        \begin{equation}
          i^* = \arg\max_{i \in \mathcal{E}} I_i(\hat{\theta}) =
            \arg\max_{i \in \mathcal{E}}
            \frac{a_i^2 \cdot P_i(\hat{\theta})(1 - P_i(\hat{\theta}))}
            {\left(\dfrac{P_i(\hat{\theta}) - c_i}{1 - c_i}\right)^2}
          \label{eq:selection}
        \end{equation}
        where $\mathcal{E}$ is the eligible item pool after steps 1--4.
  \item \textbf{Fallback}: if the eligible pool is exhausted
        (\texttt{variant\_pool\_exhausted} event emitted), the engine falls
        back to selecting the highest-information item from the full
        unrestricted pool for that KC.
\end{enumerate}

\begin{figure}[H]
\centering
\begin{tikzpicture}[
  node distance=0.9cm,
  box/.style={rectangle, draw=black!70, rounded corners=3pt, text width=5.2cm,
    align=center, fill=blue!6, font=\small, minimum height=0.75cm},
  dec/.style={diamond, draw=black!70, fill=orange!10, text width=2.8cm,
    align=center, font=\footnotesize, inner sep=1pt, aspect=2},
  arr/.style={-{Latex[length=2mm]}, thick},
  yes/.style={font=\footnotesize, left},
  no/.style={font=\footnotesize, right}
]
\node[box] (start) {Get learner state $(\hat{\theta}, p(L_n)_{1..13})$};
\node[dec, below=of start] (mastery) {Prerequisite\\KC mastered?};
\node[box, right=2.8cm of mastery] (skip) {Exclude KC from pool};
\node[dec, below=of mastery] (bloom) {Bloom level\\unlocked?};
\node[box, right=2.8cm of bloom] (skipb) {Exclude item};
\node[dec, below=of bloom] (dedup) {Variant seen\\before?};
\node[box, right=2.8cm of dedup] (skipd) {Exclude variant};
\node[box, below=of dedup] (decay) {Apply $1.5\times$ boost if KC decayed};
\node[box, below=of decay] (fisher) {Select $i^* = \arg\max\; I_i(\hat{\theta})$};
\node[box, below=of fisher] (serve) {Serve question to learner};

\draw[arr] (start) -- (mastery);
\draw[arr] (mastery) -- node[yes]{Yes} (bloom);
\draw[arr] (mastery) -- node[no]{No} (skip);
\draw[arr] (bloom) -- node[yes]{Yes} (dedup);
\draw[arr] (bloom) -- node[no]{No} (skipb);
\draw[arr] (dedup) -- node[yes]{No} (decay);
\draw[arr] (dedup) -- node[no]{Yes} (skipd);
\draw[arr] (decay) -- (fisher);
\draw[arr] (fisher) -- (serve);
\end{tikzpicture}
\caption{Adaptive question selection flowchart. Steps execute in sequence per
question request; ``Yes'' labels indicate the branch taken when the condition
is satisfied.}
\label{fig:selection_flowchart}
\end{figure}

% ─────────────────────────────────────────────────────────────────────────────
\section{IRT Calibration Pipeline}
\label{subsec:meth_calibration}
% ─────────────────────────────────────────────────────────────────────────────

IRT item parameters cannot be estimated from live session data without
sufficient observations (minimum 200 responses per item for stable 3PL
recovery \parencite{lord1980}). This project implements a two-stage offline
calibration pipeline:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{Pilot data collection.} A pilot study provides initial
        response data. The response matrix (participants $\times$ items) is
        exported from Supabase to a CSV file.
  \item \textbf{R calibration.} The \texttt{mirt} package
        \parencite{chalmers2012} fits the 3PL model with $c_i$ constrained
        to $0.25$ via an equality constraint vector. Convergence is assessed
        using RMSEA (acceptable: $< 0.08$) and the M2 statistic. If 3PL
        estimation is unstable (non-convergence after 500 EM cycles, or
        RMSEA $> 0.10$), the model falls back to Rasch (1PL). This decision
        is pre-specified as a risk mitigation and documented in the risk
        register (Table~\ref{tab:risk_register}).
  \item \textbf{JSON serialisation.} Estimated $a_i$ and $b_i$ parameters are
        exported to \texttt{question\_bank.json} and committed to the
        repository. The Node.js runtime loads this file at application startup;
        no database query is required per session.
  \item \textbf{Recalibration trigger.} If the cumulative response count for
        any item exceeds 200, a recalibration is scheduled for the next
        deployment cycle. This is tracked via a Supabase monitoring query
        against the \texttt{AnalyticsEvent} table.
\end{enumerate}

% ─────────────────────────────────────────────────────────────────────────────
\section{Pre-Test Seeding}
\label{sec:meth_pretest}
% ─────────────────────────────────────────────────────────────────────────────

Before the adaptive session begins, a 13-item diagnostic pre-test spanning all
KCs at Bloom Level 1 provides initial ability evidence, addressing the
cold-start problem \parencite{matos2025}. Items are selected at Level 1 to
minimise frustration and to ensure the pre-test is informative across the full
ability range. The raw pre-test score is converted to a Laplace-smoothed
logit seed estimate:

\begin{equation}
  \theta_{\text{seed}} = \frac{\ln\!\left(\dfrac{n_{\text{correct}} + 0.5}
    {n_{\text{incorrect}} + 0.5}\right)}{1.7}
  \label{eq:theta_seed_meth}
\end{equation}

where the $1.7$ denominator is the standard IRT logistic approximation scaling
log-odds to the normal-ogive $D$ parameter \parencite{lord1980}. Laplace
smoothing (adding $0.5$ to numerator and denominator) prevents boundary
estimates at perfect or zero pre-test scores. The seed is bounded to
$\theta_{\text{seed}} \in [-2.5, 2.5]$ and used as the initial $\hat{\theta}$
for EAP estimation in the adaptive session, with the prior standard deviation
reduced to $0.35$ to reflect the additional information.

% ─────────────────────────────────────────────────────────────────────────────
\section{Scaffolding Design}
\label{sec:meth_scaffolding}
% ─────────────────────────────────────────────────────────────────────────────

Each question item provides three levels of graduated scaffolding, accessible
on demand by the learner without interrupting the session flow. The hint
architecture follows the three-level model of \textcite{koedinger2007}:

\begin{description}
  \item[Level 1: Conceptual reframe.] Restates the question in simpler
    terms, clarifies terminology, or draws attention to relevant factual
    knowledge. Does not disclose the solution pathway.
  \item[Level 2: Procedural strategy.] Provides the method or framework
    for approaching the question. Identifies which geographic concept is
    tested without specifying the answer.
  \item[Level 3: Worked example.] Provides a fully worked analogy question
    with an explained solution. The learner's own question remains unsolved,
    but the cognitive pathway to the answer is made explicit.
\end{description}

Each hint access reduces the credit factor applied to a subsequent correct
answer (Table~\ref{tab:credit_factors}), implementing the principle that
hint-assisted correctness is less evidentially informative about latent
knowledge than unaided correctness \parencite{koedinger2007}. Incorrect
responses following any hint level are treated as full-weight incorrect
responses in the BKT update.

% ─────────────────────────────────────────────────────────────────────────────
\section{Isomorphic Variation and Cross-Session Deduplication}
\label{sec:meth_variants}
% ─────────────────────────────────────────────────────────────────────────────

A persistent methodological challenge in adaptive assessment is item
overexposure: learners who return for multiple sessions may receive previously
seen questions, inflating mastery estimates. This project addresses the
problem through a two-layer variant system.

\textbf{Isomorphic variants.} Each KC-Bloom-level combination is authored as a
\texttt{QuestionTemplate} with a stem and a set of correct/distractor answer
keys. Multiple \texttt{QuestionVariant} records are derived from each template
by substituting surface-level details (specific country names, values,
locations) while preserving the underlying cognitive demand. Isomorphic
variants test the same KC-Bloom construct as their parent template but differ
in surface form, rendering rote memorisation of prior answers non-transferable.

\textbf{Cross-session deduplication.} Every variant served is recorded in the
\texttt{UserVariantSeen} join table, keyed to the user and variant IDs. At
question selection, the eligible pool excludes any variant present in this
table for the requesting user. If all variants of a template have been
exhausted, the \texttt{variant\_pool\_exhausted} analytics event is emitted
and the engine falls back to the template-level item pool.

% ─────────────────────────────────────────────────────────────────────────────
\section{Evaluation Protocol}
\label{sec:meth_evaluation}
% ─────────────────────────────────────────────────────────────────────────────

\subsection{Study Design}
\label{subsec:meth_study}

The evaluation employs a within-subjects quasi-experimental design with the
following conditions:

\begin{description}
  \item[Adaptive condition.] Full adaptive session using the IRT+BKT engine
    with Bloom-gated selection, pre-test seeding, and graduated scaffolding.
  \item[Baseline reference.] VanLehn's (2011) meta-analytic benchmark of
    $d = 0.76$ for step-based ITS systems is used as the primary comparison
    standard, as a direct no-adaptation control arm could not be implemented
    within the single-session constraint.
\end{description}

\subsection{Participants}
\label{subsec:meth_participants}

Participants are recruited from University of Hull undergraduate and
postgraduate students via the university's participant pool and departmental
email lists. Inclusion criteria: (a) English-language fluency; (b) no
prior formal study of UK KS3/GCSE Geography at GCSE level or above (to avoid
ceiling effects); (c) informed written consent. Target sample size is $n \geq
15$ for the primary session, with a retention follow-up targeting $n \geq 10$
after anticipated attrition.

\subsection{Instruments}
\label{subsec:meth_instruments}

\begin{description}
  \item[System Usability Scale (SUS).] The 10-item SUS questionnaire
    \parencite{brooke1996} is administered at session end. The benchmark for
    above-average usability is $\geq 68$ \parencite{bangor2008}.
  \item[Misconception analysis.] Post-session distractor frequency tables
    identify the most common incorrect attractors per KC, enabling qualitative
    analysis of systematic knowledge gaps.
  \item[Theta trajectory.] Session-level $\hat{\theta}$ sequences are stored
    for longitudinal analysis of within-session learning gains.
  \item[7-day retention test.] A subset of isomorphic items (not previously
    seen) is administered via the same platform seven days post-session to
    assess delayed retention of mastered KCs.
\end{description}

\subsection{Limitations of the Evaluation}
\label{subsec:meth_limitations}

The evaluation design carries several acknowledged limitations:

\begin{itemize}[leftmargin=2em]
  \item \textbf{Sample size.} $n = 4$ external participants were available for
        the primary evaluation phase, substantially below the target of $n = 15$.
        This precludes inferential statistical testing (paired $t$-test requires
        approximately $n = 13$ for 80\% power at $d = 0.76$, $\alpha = 0.05$).
        Results are reported descriptively as proof-of-concept evidence only.
  \item \textbf{No control condition completed.} The within-subjects control arm
        was not implemented due to time constraints; VanLehn's benchmark serves
        as the comparative reference.
  \item \textbf{Convenience sample.} University students are not representative
        of the KS3/GCSE student population the system targets. Findings cannot
        be generalised to secondary school learners without further validation.
  \item \textbf{IRT parameters from pilot.} Item parameters are estimated from
        pilot data with a small calibration sample. Parameter instability is
        a known risk and is mitigated through the constrained calibration
        strategy described in Section~\ref{subsec:meth_calibration}.
\end{itemize}

% ─────────────────────────────────────────────────────────────────────────────
\section{System Architecture Overview}
\label{sec:meth_architecture}
% ─────────────────────────────────────────────────────────────────────────────

\begin{figure}[H]
\centering
\begin{tikzpicture}[
  node distance=1.2cm and 1.8cm,
  layer/.style={rectangle, draw=black!60, rounded corners=4pt,
    fill=gray!8, text width=3.0cm, align=center,
    font=\small\bfseries, minimum height=1.0cm},
  mod/.style={rectangle, draw=black!50, rounded corners=3pt,
    fill=blue!7, text width=2.6cm, align=center,
    font=\footnotesize, minimum height=0.8cm},
  db/.style={cylinder, shape border rotate=90, draw=black!60,
    fill=green!8, text width=2.2cm, align=center,
    font=\footnotesize, minimum height=1.0cm, aspect=0.25},
  arr/.style={-{Latex[length=2mm]}, thick, gray!70}
]
% Layers
\node[layer] (client) {Browser\\Client};
\node[layer, below=of client] (nextjs) {Next.js\\App Router};
\node[layer, below=of nextjs] (api) {API Routes\\(Edge Functions)};
\node[layer, below=of api] (engine) {Adaptive Engine};
\node[db, below=of engine] (db) {Supabase\\PostgreSQL};

% Engine sub-modules
\node[mod, right=2.0cm of engine] (irt) {IRT 3PL\\$\hat{\theta}$ EAP};
\node[mod, above=0.5cm of irt] (bkt) {BKT\\per-KC state};
\node[mod, below=0.5cm of irt] (sel) {Item\\Selector};

% Arrows
\draw[arr] (client) -- node[right, font=\footnotesize]{HTTPS} (nextjs);
\draw[arr] (nextjs) -- (api);
\draw[arr] (api) -- (engine);
\draw[arr] (engine) -- (db);
\draw[arr] (engine) -- (irt);
\draw[arr] (engine) -- (bkt);
\draw[arr] (engine) -- (sel);
\draw[arr, dashed] (db) to[bend right=30] node[left, font=\footnotesize]{Prisma ORM} (api);
\end{tikzpicture}
\caption{High-level system architecture. The adaptive engine is isolated as a
pure-function module (\texttt{src/lib/algorithms/}) with no direct database
access; all I/O passes through the API route layer.}
\label{fig:architecture}
\end{figure}

The system is deployed as a Next.js 14 (App Router) application on Vercel,
with Supabase (Frankfurt region) providing the PostgreSQL database, Row-Level
Security policies, and authentication services. Prisma ORM manages schema
migrations and type-safe database access. The adaptive engine module
(\texttt{src/lib/algorithms/}) is intentionally isolated from the database
layer: it receives serialised learner state as input and returns a question
selection decision as output, making it independently testable and
domain-agnostic.

% ─────────────────────────────────────────────────────────────────────────────
\section{Agile Development Methodology}
\label{sec:meth_agile}
% ─────────────────────────────────────────────────────────────────────────────

Development followed a two-week sprint cycle adapted from Scrum, managed
through a Notion project tracker. The sprint structure was chosen over a
waterfall approach because the adaptive algorithm design required iterative
empirical validation: initial IRT parameter estimates are hypotheses, not
specifications, and the system design required multiple calibration cycles to
stabilise. This decision aligns with the CSSE competency requirement to
``use established software development methodologies to discover authentic
requirements, respond to change, ensure quality and deliver value to users
and stakeholders'' \parencite{hull2024}.

\subsection{Sprint Structure}
\label{subsec:meth_sprints}

\begin{table}[H]
\centering
\caption{Sprint history: objectives, deliverables, and retrospective notes.}
\label{tab:sprints}
\renewcommand{\arraystretch}{1.25}
\begin{tabularx}{\textwidth}{clXX}
\toprule
\textbf{Sprint} & \textbf{Dates} & \textbf{Objective} & \textbf{Key deliverable / deviation} \\
\midrule
1 & Oct--Nov 2025 & Requirements, KC ontology, technology selection &
  13 KC prerequisite graph; technology ADRs (Next.js + Supabase) \\
2 & Nov--Dec 2025 & Core adaptive engine, database schema, pre-test seeding &
  \texttt{adaptive-engine.ts}; Prisma schema; EAP implementation \\
3 & Dec 2025--Jan 2026 & Question bank, IRT calibration pipeline, BKT &
  \texttt{seedKCs.ts} (13 KCs, 3 Bloom levels); R calibration script \\
4 & Jan--Feb 2026 & Scaffolding (hint system), variant deduplication &
  \texttt{hints/route.ts}; \texttt{UserVariantSeen} table; credit-factor BKT \\
5 & Feb 2026 & Scope reduction decision; evaluation redesign &
  Scope locked to single-KC validation; VanLehn benchmark adopted as comparator \\
6 & Feb--Mar 2026 & Session continuity, onboarding tour, pause/resume &
  \texttt{FirstVisitTour.tsx}; pause menu; session restore logic \\
7 & Mar 2026 & Security hardening (NIST 800-63B), rate limiting, RLS &
  Auth middleware; Supabase RLS policies; bcrypt migration \\
8 & Mar--Apr 2026 & Leaderboard, XP system, confidence calibration &
  \texttt{leaderboard/route.ts}; XP/level schema; decay notifications \\
9 & Apr 2026 & Evaluation data collection, SUS administration &
  SUS instrument deployed; participant sessions recorded \\
10 & Apr--May 2026 & Dissertation writing, final polishing &
  Dissertation LaTeX; this chapter \\
\bottomrule
\end{tabularx}
\end{table}

\subsection{Responding to Change: The Scope Reduction Decision}
\label{subsec:meth_scope_reduction}

The initial project proposal described a multi-domain adaptive system covering
STEM and Humanities subjects. By Sprint 5 (February 2026), technical risk
assessment revealed that multi-domain calibration would require separate IRT
parameter estimates for each domain, each demanding a minimum of 200 responses
per item, infeasible within the available study timeline. Furthermore,
deploying a validation study across multiple domains would have diluted the
evaluation's statistical power across subgroups too small for meaningful
analysis.

The decision to scope down to a single domain (UK Geography) and
subsequently to validate a single core KC workflow was taken in consultation
with the project supervisor and is documented in the project's Scope Lock
Document. This decision is an instance of the CSSE competency ``respond to
change'': it represents an evidence-based pivot from an overscoped requirement
to a tractable research claim, not a retreat from ambition. The system's
domain-agnostic architecture (the adaptive engine operates on any KC graph,
regardless of subject domain) preserves the multi-domain vision as a credible
future direction.

\subsection{Version Control and Audit Trail}
\label{subsec:meth_git}

All development activity is version-controlled in a public GitHub repository
(\url{https://github.com/dlalloyd/E-Learning-Final-Project}). The repository
provides a verifiable audit trail of development progress: each feature is
committed atomically with a descriptive commit message, enabling reconstruction
of the development timeline by examining the Git log. Key architectural
milestones are tagged in the repository to facilitate evaluation.

% ─────────────────────────────────────────────────────────────────────────────
\section{Project Timeline}
\label{sec:meth_gantt}
% ─────────────────────────────────────────────────────────────────────────────

Figure~\ref{fig:gantt} presents the planned and actual project timeline,
reconstructed from the Git commit history. The Sprint~5 scope reduction
(red dashed line) and the resulting re-planning are clearly visible.

\begin{figure}[H]
\centering
\begin{tikzpicture}[x=0.92cm, y=0.72cm]
  % ── Vertical grid
  \foreach \x in {0,...,10} {
    \draw[gray!22, thin] (\x, 0.65) -- (\x, -4.60);
  }
  % ── Sprint header
  \foreach \i in {1,...,10} {
    \pgfmathsetmacro\xc{\i - 0.5}
    \node[font=\footnotesize\bfseries] at (\xc, 0.45) {S\i};
  }
  % ── Month sub-row
  \foreach \xc/\mo in
      {0.5/Oct, 1.5/Nov, 2.5/Dec, 3.5/Jan, 4.5/Feb,
       5.5/Mar, 6.5/Mar, 7.5/Apr, 8.5/Apr, 9.5/May} {
    \node[font=\tiny, gray!60] at (\xc, 0.10) {\mo};
  }
  % ── Phase bars
  % Initiation & Design  S1–S4
  \fill[blue!28, rounded corners=2pt] (0,-0.30) rectangle (4,-0.80);
  \node[anchor=east, font=\small] at (-0.22,-0.55) {Initiation \& Design};
  % Core Development  S2–S5
  \fill[blue!48, rounded corners=2pt] (1,-1.00) rectangle (5,-1.50);
  \node[anchor=east, font=\small] at (-0.22,-1.25) {Core Development};
  % Integration & Hardening  S6–S8
  \fill[blue!68, rounded corners=2pt] (5,-1.70) rectangle (8,-2.20);
  \node[anchor=east, font=\small] at (-0.22,-1.95) {Integration \& Hardening};
  % Evaluation  S9
  \fill[orange!55, rounded corners=2pt] (8,-2.40) rectangle (9,-2.90);
  \node[anchor=east, font=\small] at (-0.22,-2.65) {Evaluation ($n=4$)};
  % Dissertation  S9–S10
  \fill[green!35!black!55, rounded corners=2pt] (8,-3.10) rectangle (10,-3.60);
  \node[anchor=east, font=\small] at (-0.22,-3.35) {Dissertation};
  % ── Scope reduction marker  x = 4
  \draw[red!65, thick, dashed] (4, 0.65) -- (4,-4.10);
  \node[red!65, font=\tiny, align=center, above] at (4, 0.65)
      {\textbf{Scope}\\reduction};
  % ── Submission diamond  x = 10
  \fill[red!60]
      (10.00,-3.35) -- (10.22,-3.15) -- (10.44,-3.35) -- (10.22,-3.55) -- cycle;
  \node[anchor=west, font=\tiny, red!70] at (10.50,-3.35) {06~May};
  % ── Border
  \draw[black!25] (0, 0.65) rectangle (10,-4.10);
\end{tikzpicture}
\caption{Project Gantt chart. Each column = one fortnightly sprint (Oct~2025–May~2026).
Bar shading indicates development phase; the red dashed line marks the Sprint~5
scope reduction from multi-domain to single-domain Geography KC validation.}
\label{fig:gantt}
\end{figure}

% ─────────────────────────────────────────────────────────────────────────────
\section{Risk Register}
\label{sec:meth_risks}
% ─────────────────────────────────────────────────────────────────────────────

\begin{table}[H]
\centering
\caption{Project risk register. Likelihood: H = High, M = Medium, L = Low.
Impact: H = High, M = Medium, L = Low.}
\label{tab:risk_register}
\renewcommand{\arraystretch}{1.25}
\begin{tabularx}{\textwidth}{cp{4.5cm}ccp{2cm}X}
\toprule
\textbf{ID} & \textbf{Risk} & \textbf{L} & \textbf{I} & \textbf{Status} & \textbf{Mitigation} \\
\midrule
R1 & IRT 3PL fails to converge (small pilot sample) &
  H & M & Mitigated &
  Pre-specify Rasch (1PL) fallback; fix $c=0.25$ as equality constraint \\
R2 & Participant attrition at 7-day follow-up &
  H & M & Accepted &
  Target $n=18$ recruitment; acknowledge in evaluation limitations \\
R3 & variant\_pool\_exhausted before session end &
  M & L & Mitigated &
  Fallback to unrestricted pool; \texttt{variant\_pool\_exhausted} event emitted \\
R4 & Supabase RLS misconfiguration exposing user data &
  L & H & Mitigated &
  RLS policies peer-reviewed; rate limiting on auth endpoints \\
R5 & Cross-session BKT state corruption (concurrent sessions) &
  L & H & Mitigated &
  Postgres row-level locking; session state written atomically \\
R6 & Scope overrun (multi-domain ambition) &
  H & H & Resolved &
  Sprint 5 scope reduction to single-domain Geography KC validation \\
R7 & Gantt chart absent from interim report &
  H & M & Resolved &
  Gantt included in this final report (Figure~\ref{fig:gantt}) \\
\bottomrule
\end{tabularx}
\end{table}

% ─────────────────────────────────────────────────────────────────────────────
\section{What Could and Could Not Be Delivered}
\label{sec:meth_delivery}
% ─────────────────────────────────────────────────────────────────────────────

\subsection{Delivered}
\label{subsec:meth_delivered}

The following components were fully implemented and are verifiable in the
production deployment and Git history:

\begin{itemize}[leftmargin=2em]
  \item Hybrid IRT 3PL + BKT adaptive engine with EAP estimation,
        Bloom-gated selection, and credit-factor BKT modification
  \item 13-KC question bank with isomorphic variants and cross-session
        deduplication (\texttt{UserVariantSeen} table)
  \item Three-level graduated hint system with evidential credit weighting
  \item Pre-test seeding with Laplace-smoothed logit initialisation
  \item BKT decay detection and successive relearning boost ($1.5\times$)
  \item Five-tab session summary dashboard (Trajectory, Skills, Misconceptions,
        Achievements, Share)
  \item Leaderboard with XP ranking, consent opt-in, and rank-change animation
  \item First-visit onboarding tour (\texttt{FirstVisitTour.tsx})
  \item Security implementation: bcrypt, JWT, rate limiting, Supabase RLS,
        NIST SP 800-63B compliance
  \item Production deployment on Vercel with Supabase Frankfurt backend
\end{itemize}

\subsection{Not Delivered}
\label{subsec:meth_not_delivered}

The following items from the original proposal were not delivered, with
justifications:

\begin{itemize}[leftmargin=2em]
  \item \textbf{Multi-domain support.} Evaluated and superseded by the scope
        reduction decision (Section~\ref{subsec:meth_scope_reduction}). The
        architecture supports multi-domain extension without algorithmic change.
  \item \textbf{K-means cold-start clustering.} Originally proposed as an
        alternative cold-start mechanism. Evaluated and superseded by the
        pre-test seeding approach, which provides per-KC ability evidence
        rather than a demographic cluster assignment. Pre-test seeding is
        both more principled and more interpretable.
  \item \textbf{Full A/B randomised control trial.} A direct no-adaptation
        control arm was not implemented within the single-session study
        design; VanLehn's meta-analytic benchmark is used as the comparator.
  \item \textbf{7-day retention data at scale.} Follow-up data collection
        could not be completed within the project timeline; retention
        assessment was conducted for the available $n=4$ sample only.
\end{itemize}

% ─────────────────────────────────────────────────────────────────────────────
\section{Quality Assurance}
\label{sec:meth_qa}
% ─────────────────────────────────────────────────────────────────────────────

Quality assurance mechanisms employed across the project lifecycle include:

\begin{itemize}[leftmargin=2em]
  \item \textbf{TypeScript strict mode.} The entire codebase compiles under
        TypeScript strict mode, eliminating a class of runtime errors at
        the type-checking stage.
  \item \textbf{Prisma schema validation.} Database schema changes are
        managed exclusively through Prisma migration files, providing a
        versioned, auditable schema history.
  \item \textbf{Vercel preview deployments.} Each feature branch receives an
        isolated preview deployment, enabling integration testing before
        merge to main.
  \item \textbf{Supabase RLS verification.} Row-Level Security policies were
        verified by attempting cross-user data access from a test account,
        confirming that users cannot read each other's session data.
  \item \textbf{Manual SUS administration.} SUS instrument was administered
        by the researcher, with participants reading and signing a
        participant information sheet confirming informed consent before
        beginning any session.
\end{itemize}
