\chapter{Implementation}
\label{ch:implementation}

% ─────────────────────────────────────────────────────────────────────────────
\section{System Architecture}
\label{sec:impl_architecture}
% ─────────────────────────────────────────────────────────────────────────────

GeoMentor is implemented as a monolithic Next.js 14 application using the App
Router paradigm, deployed to Vercel's edge network with a Supabase PostgreSQL
database hosted in the Frankfurt (eu-central-1) region. This architectural
choice departs from a microservices pattern: the adaptive engine, API routes,
and frontend components all reside in a single codebase. This was a deliberate
CSSE decision — the system's research contribution lies in the adaptive
algorithm, not in distributed systems complexity, and a monolith reduced
operational overhead and eliminated the inter-service latency that would
have degraded the interactive learning experience.

The key architectural constraint is the \textbf{isolation of the adaptive
engine}: the algorithm modules (\texttt{src/lib/algorithms/irt.ts} and
\texttt{src/lib/algorithms/bkt.ts}) are implemented as pure functions with no
direct database access. All database I/O passes through Next.js API route
handlers, which orchestrate the learner state lifecycle — loading state,
invoking engine functions, persisting updated state, and returning the
question to the client. This isolation makes the engine independently testable
and domain-agnostic: replacing the Geography question bank with any other KC
graph requires no changes to the algorithm modules.

\subsection{Technology Stack}
\label{subsec:impl_stack}

\begin{table}[H]
\centering
\caption{Technology stack with architectural justification.}
\label{tab:tech_stack}
\renewcommand{\arraystretch}{1.25}
\begin{tabularx}{\textwidth}{llX}
\toprule
\textbf{Layer} & \textbf{Technology} & \textbf{Justification} \\
\midrule
Frontend & Next.js 14 (App Router) &
  React Server Components reduce client bundle size; streaming enables
  progressive loading of session summary data without a full re-render \\
Styling & Tailwind CSS &
  Utility-first CSS eliminates style-sheet maintenance overhead;
  consistent design tokens across all components \\
ORM & Prisma 5 &
  Type-safe database access; migration history provides schema audit trail
  (CSSE quality assurance requirement) \\
Database & Supabase PostgreSQL &
  Row-Level Security (RLS) enforces data isolation at the database layer,
  not the application layer; reduces risk of privilege escalation bugs \\
Auth & Supabase Auth + JWT &
  PKCE flow; bcrypt password hashing; NIST SP 800-63B compliant \\
Deployment & Vercel &
  Preview deployments per branch; automatic HTTPS; edge caching \\
IRT calibration & R + \texttt{mirt} &
  Industry-standard psychometric tooling; calibration is offline and
  audit-logged in the repository \\
\bottomrule
\end{tabularx}
\end{table}

% ─────────────────────────────────────────────────────────────────────────────
\subsection{Non-Functional Properties}
\label{subsec:impl_nfr}

The architecture was designed against the following non-functional targets,
relevant to the CP6 CSSE competency requirement:

\begin{description}
  \item[Scalability.] The monolithic Next.js deployment scales horizontally
    via Vercel's serverless function infrastructure: each API route invocation
    is an independent function execution, and Vercel automatically provisions
    additional instances under load. The primary scalability constraint is
    the PostgreSQL connection pool; Supabase's PgBouncer connection pooler
    mitigates this for the expected evaluation-scale traffic. For a deployment
    beyond a few hundred concurrent learners, connection pooling configuration
    and read-replica routing would be required.

  \item[Reliability.] The system targets \textgreater\,99\% uptime via
    Vercel's edge network SLA. Supabase Frankfurt operates on a managed
    PostgreSQL cluster with daily automated backups and point-in-time recovery.
    Error boundaries (\texttt{error.tsx} boundary components) prevent
    component-level failures from crashing the full session.

  \item[Efficiency.] The adaptive engine executes the EAP grid integration
    and BKT update in-process within the API route, targeting an end-to-end
    question selection latency of under 500\,ms (measured from request receipt
    to response delivery). The JSON-serialised \texttt{kcStates} field on the
    \texttt{Session} model was chosen to eliminate join overhead on the
    hot path.

  \item[Maintainability.] TypeScript strict mode enforces type safety at
    every layer; Prisma migration files provide a versioned, auditable schema
    history; and Git atomic commits with descriptive messages enable precise
    reconstruction of the development timeline. The adaptive engine's
    pure-function architecture means algorithm changes can be made without
    touching the database or API layers.
\end{description}

% ─────────────────────────────────────────────────────────────────────────────
\section{Database Schema}
\label{sec:impl_schema}
% ─────────────────────────────────────────────────────────────────────────────

The Prisma schema defines 20 models. The six models central to the adaptive
engine are described here; the full schema is available in the repository
(\texttt{prisma/schema.prisma}).

\subsection{Session Model}
\label{subsec:impl_session}

The \texttt{Session} model is the principal state carrier for the adaptive
engine. Key fields:

\begin{itemize}[leftmargin=2em]
  \item \texttt{theta~Float} — current EAP ability estimate, initialised at
        $-0.780$ (the 20th percentile derived from the self-pilot calibration,
        2026-02-26).
  \item \texttt{thetaSd~Float} — posterior standard deviation, initialised at
        $0.543$, used to widen or tighten the EAP prior.
  \item \texttt{kcStates~String} — JSON-serialised
        \texttt{Record\textless{}kcId, KCState\textgreater{}}, carrying the
        full BKT state for all 13 Knowledge Components between requests.
        JSON serialisation was chosen over a normalised \texttt{KCState} table
        to reduce read latency: a single \texttt{Session.findUnique} call
        returns all per-KC state without joins.
  \item \texttt{condition~String} — \texttt{"adaptive"} or \texttt{"static"},
        assigned at signup via the \texttt{studyCondition} field on the
        \texttt{User} model, preventing self-selection bias in the
        within-subjects evaluation design.
  \item \texttt{startedFromAssessmentId~String?} — links to the pre-test
        \texttt{Assessment} record that seeded the session's initial
        $\hat{\theta}$, providing a verifiable audit trail of cold-start
        seeding decisions.
\end{itemize}

\subsection{Interaction Model}
\label{subsec:impl_interaction}

The \texttt{Interaction} model records every question--answer event at full
psychometric resolution:

\begin{itemize}[leftmargin=2em]
  \item \texttt{theta\_before}, \texttt{theta\_after} — EAP estimates
        immediately before and after the response, enabling per-item
        $\Delta\theta$ analysis.
  \item \texttt{pLearned\_before}, \texttt{pLearned\_after} — BKT posteriors
        for the question's KC, enabling per-response knowledge state analysis.
  \item \texttt{hintsUsed~Int}, \texttt{hintLevelMax~Int} — total hints
        accessed and the maximum hint level revealed, used by the credit
        factor computation in the BKT update.
  \item \texttt{cognitiveLoad~Int?} — optional NASA-TLX style self-report
        (1=Very Easy, 5=Very Difficult) collected after each question,
        enabling cognitive load analysis across Bloom levels.
  \item \texttt{responseTimeMs~Int} — response latency in milliseconds,
        providing a behavioural correlate of item difficulty.
\end{itemize}

\subsection{Question Bank Models}
\label{subsec:impl_question_bank}

The question bank is organised as a two-layer hierarchy:

\begin{description}
  \item[\texttt{QuestionTemplate}] defines the KC, Bloom level, IRT
    parameters ($a$, $b$, $c$), and a stem template containing
    \texttt{\{variable\}} placeholders.
  \item[\texttt{QuestionVariant}] substitutes specific values for each
    placeholder, producing a surface-different isomorphic item. Multiple
    variants per template prevent item overexposure across sessions.
  \item[\texttt{UserVariantSeen}] records every (user, variant) pair that
    has been presented, enforced as a unique constraint. At question
    selection, the eligible pool excludes any variant present in this table
    for the requesting user.
  \item[\texttt{VariantPresentation}] logs each presentation event with
    timestamp and session context, supporting post-hoc analysis of
    item exposure patterns.
\end{description}

\subsection{Analytics Model}
\label{subsec:impl_analytics}

The \texttt{AnalyticsEvent} model provides a structured event log for the
five-track evaluation (Chapter~\ref{ch:results}). Events of interest include:

\begin{itemize}[leftmargin=2em]
  \item \texttt{bloom\_escalation} — emitted when a learner's Bloom gate
        advances, with payload recording the KC, new Bloom level, and
        triggering $p(\text{Learned})$ value.
  \item \texttt{variant\_pool\_exhausted} — emitted when no unseen variants
        remain for a KC-Bloom combination, triggering the fallback selection
        logic.
  \item \texttt{kc\_mastered} — emitted when $p(\text{Learned}) \geq 0.95$
        for a KC, with payload recording the number of questions required.
  \item \texttt{decay\_detected} — emitted at session start when a KC's
        cross-session $p(\text{Learned})$ decline exceeds the decay threshold.
\end{itemize}

% ─────────────────────────────────────────────────────────────────────────────
\section{Adaptive Engine Implementation}
\label{sec:impl_engine}
% ─────────────────────────────────────────────────────────────────────────────

\subsection{IRT Module (\texttt{src/lib/algorithms/irt.ts})}
\label{subsec:impl_irt}

The IRT module implements the 3PL model with the $D = 1.7$ logistic
approximation \parencite{lord1980}. The core probability function is:

\begin{lstlisting}[language=JavaScript, caption={3PL probability function
(\texttt{irt.ts:91--94}).}, label=lst:p3pl]
export function p3PL(theta: number, params: IRTParams): number {
  const { a, b, c } = params;
  return c + (1 - c) / (1 + Math.exp(-D * a * (theta - b)));
}
\end{lstlisting}

The Fisher information function implements Lord's (1980) Eq.~6.19 for the
3PL model:

\begin{lstlisting}[language=JavaScript, caption={Fisher information function
(\texttt{irt.ts:100--105}).}, label=lst:fisher]
export function itemInformation(theta: number, params: IRTParams): number {
  const { a, c } = params;
  const p = p3PL(theta, params);
  return (D ** 2 * a ** 2 * ((p - c) ** 2)) / ((1 - c) ** 2 * p * (1 - p));
}
\end{lstlisting}

The EAP estimator integrates over a grid of $K = 161$ points spanning
$\theta \in [-4, 4]$ with step $0.05$. The finer grid (compared to the
$0.1$-step grid proposed in the design phase) was adopted after self-pilot
testing revealed that coarser grids produced visible ``staircase'' artefacts
in the $\hat{\theta}$ trajectory for high-ability learners. The prior is a
normal distribution centred at $\mu_0 = -0.780$ with $\sigma_0 = 0.543$,
derived from the developer's own self-pilot session on 2026-02-26, which
provided empirical grounding for the novice prior rather than an arbitrary
assumed value.

\subsection{BKT Module (\texttt{src/lib/algorithms/bkt.ts})}
\label{subsec:impl_bkt}

The BKT module implements the standard Corbett--Anderson update plus the
credit factor modification. The implementation adopts a \textit{soft
observation} interpolation for credit-weighted correct responses, mixing the
Bayesian posterior for a full correct answer with the prior in proportion to
the credit factor:

\begin{lstlisting}[language=JavaScript, caption={Credit factor implementation
in \texttt{updateBKT()} (\texttt{bkt.ts:159--165}).}, label=lst:credit]
if (isCorrect) {
  pLearnedGivenResponse =
    factor * posteriorIfCorrect + (1 - factor) * pLearned_before;
} else {
  pLearnedGivenResponse = posteriorIfIncorrect;
}
\end{lstlisting}

This soft-observation approach is mathematically equivalent to a Bayesian
update with a degraded likelihood, but is numerically more stable than
computing a partial likelihood directly. Incorrect responses following hint
access receive full-weight Bayesian updates (no credit factor applied):
providing a hint does not make an incorrect response less informative about the
learner's knowledge state.

\textbf{Mastery threshold.} The BKT mastery threshold is implemented at
$p(\text{Learned}) \geq 0.95$, higher than the $0.80$ threshold cited in the
literature review. This decision reflects the binary nature of the BKT latent
state: because BKT models a hard ``knows / does not know'' distinction, a
higher threshold provides a stronger guarantee that the posterior is not merely
elevated by a lucky guess run. The $0.80$ threshold applies to the IRT-based
mastery check (\texttt{checkMastery()} in \texttt{irt.ts}), which averages
$P(\text{correct})$ across all KC items at the current $\hat{\theta}$ and uses
the continuous IRT scale rather than a binary state estimate. The two thresholds
address different aspects of mastery: BKT monitors \textit{knowledge state
transitions}; IRT monitors \textit{average response probability}.

The KC parameter set for the deployed system reflects the Geography KC
structure described in Table~\ref{tab:kc_table}, with parameters seeded from
curriculum difficulty estimates and the DfE Geography specification
\parencite{dfe2013}.

% ─────────────────────────────────────────────────────────────────────────────
\section{API Route Design}
\label{sec:impl_api}
% ─────────────────────────────────────────────────────────────────────────────

The adaptive session lifecycle is managed by four API routes:

\begin{description}
  \item[\texttt{GET /api/sessions/[id]/next-question}]
    The core adaptive selection route. Execution sequence:
    \begin{enumerate}[leftmargin=2em, nosep]
      \item Load session (theta, thetaSd, kcStates, condition).
      \item Load unseen variant pool: join \texttt{QuestionTemplate}
            $\times$ \texttt{QuestionVariant} minus
            \texttt{UserVariantSeen} for this user.
      \item Apply mastery gate (exclude templates whose KC has unmastered
            prerequisites in the prerequisite DAG).
      \item Apply Bloom gate (exclude templates whose Bloom level exceeds
            the learner's current escalation threshold for that KC).
      \item Apply decay boost ($1.5\times$ information multiplier) for KCs
            flagged with cross-session $p(\text{Learned})$ decline.
      \item Select $i^* = \arg\max_i I_i(\hat{\theta})$ from eligible pool.
      \item Render variant: interpolate \texttt{\{variable\}} placeholders
            with variant-specific values; shuffle answer options using
            Fisher-Yates to prevent position bias.
      \item Emit \texttt{variant\_pool\_exhausted} if eligible pool was
            empty (fallback to unrestricted selection).
      \item Return serialised question object.
    \end{enumerate}

  \item[\texttt{POST /api/sessions/[id]/answer}]
    Processes a learner response and updates the learner state:
    \begin{enumerate}[leftmargin=2em, nosep]
      \item Validate the submitted answer label against the stored correct
            label for the presented variant.
      \item Compute credit factor from \texttt{hintLevelMax}.
      \item Run EAP update over the full response history to produce new
            $\hat{\theta}$ and $\sigma_\theta$.
      \item Run BKT update for the question's KC with the credit factor.
      \item Persist updated session state (theta, thetaSd, kcStates).
      \item Write \texttt{Interaction} record with full psychometric payload.
      \item Check mastery: if BKT posterior $\geq 0.95$ and not previously
            mastered, emit \texttt{kc\_mastered} event, increment
            \texttt{User.kcsMastered}, award XP.
      \item Check Bloom escalation: if $p(\text{Learned})$ has crossed an
            escalation threshold, emit \texttt{bloom\_escalation} event.
    \end{enumerate}

  \item[\texttt{GET /api/sessions/[id]/summary}]
    Generates the session summary payload for the five-tab dashboard.
    Aggregates: per-KC BKT trajectory, overall $\hat{\theta}$ delta,
    distractor frequency table, XP earned, and a natural-language narrative
    summarising session performance.

  \item[\texttt{GET /api/sessions/[id]/trajectory}]
    Returns the longitudinal $\hat{\theta}$ sequence across all interactions
    in the session, used by the Trajectory tab of the session summary
    dashboard to render the ability progression chart.
\end{description}

An additional route, \texttt{GET /api/analytics/error-patterns}, aggregates
distractor selection frequencies across all interactions for a given KC,
returning the most common incorrect attractors. This route powers the
Misconceptions tab and constitutes Track 4 of the multi-method evaluation
(Section~\ref{sec:results_misconceptions}).

% ─────────────────────────────────────────────────────────────────────────────
\section{Security Implementation}
\label{sec:impl_security}
% ─────────────────────────────────────────────────────────────────────────────

Security requirements were specified against NIST Special Publication 800-63B
\parencite{nist2017}. The following controls were implemented:

\begin{description}
  \item[Password storage.] Passwords are hashed with bcrypt at cost factor 12
    before storage. Plaintext passwords are never persisted or logged.
  \item[Authentication.] JWT-based session tokens with Supabase Auth PKCE
    flow. Tokens are short-lived (1 hour) with automatic refresh. Token
    payloads contain only the user ID and role; no personal data is encoded
    in the token.
  \item[Rate limiting.] Authentication endpoints (login, registration,
    password reset) are rate-limited to 10 requests per minute per IP address,
    mitigating credential stuffing and brute-force attacks.
  \item[Row-Level Security.] Supabase RLS policies enforce that learners can
    only read and write their own \texttt{Session}, \texttt{Interaction}, and
    \texttt{Assessment} records. Policies were verified by cross-user access
    testing from a separate test account.
  \item[Email blocklist.] Disposable email domain registration is blocked at
    signup via a maintained blocklist, reducing research data pollution from
    low-effort registrations.
  \item[Input validation.] All API route inputs are validated server-side
    against typed schemas before processing. Client-side validation is
    supplementary only.
  \item[HTTPS.] All traffic is served over HTTPS via Vercel's edge
    certificates. HTTP requests are redirected at the CDN layer.
\end{description}

% ─────────────────────────────────────────────────────────────────────────────
\section{Frontend Components}
\label{sec:impl_frontend}
% ─────────────────────────────────────────────────────────────────────────────

\subsection{Session Flow and Question Interface}
\label{subsec:impl_session_flow}

The learner session follows a linear state machine managed by the top-level
session page component. States are: \texttt{pre-test} $\to$
\texttt{instruction} $\to$ \texttt{question} $\to$ \texttt{feedback} $\to$
\texttt{summary}. The question interface renders:

\begin{itemize}[leftmargin=2em]
  \item The question stem and four labelled answer options (A--D), shuffled
        per-presentation to prevent position-bias memorisation.
  \item A progress bar showing overall session completion and the current KC's
        BKT progress toward mastery.
  \item A three-level hint panel (\texttt{HintPanel.tsx}), accessible on
        demand. Each hint level is revealed progressively; previously revealed
        levels remain visible. The panel records the maximum hint level
        accessed before the learner submits their answer.
  \item A confidence slider (1--5) presented before answer submission,
        enabling the confidence calibration analysis reported in
        Track 4 of the evaluation.
  \item A pause button that persists the current session state to the database
        and allows the learner to resume from the exact same question in a
        future visit.
\end{itemize}

\subsection{First-Visit Onboarding Tour}
\label{subsec:impl_tour}

\texttt{FirstVisitTour.tsx} implements a five-step onboarding overlay
displayed to new users on their first session. Steps explain: (1) the
adaptive engine's purpose, (2) the BKT mastery bar and what it means,
(3) how the hint system works and its effect on mastery estimates, (4) the
XP and level system, and (5) the session summary dashboard. Tour completion
is persisted to \texttt{localStorage} to prevent repetition. This implements
the signalling principle from Mayer's (2009) cognitive theory of multimedia
learning: the adaptive system's reasoning is made legible before the learner
encounters the first question.

\subsection{Session Summary Dashboard}
\label{subsec:impl_summary}

\texttt{SessionSummaryDashboard.tsx} presents a five-tab post-session
analysis interface:

\begin{description}
  \item[Overview.] Overall session statistics: total questions, accuracy,
    XP earned, level achieved, and a natural-language performance narrative
    generated by the summary API route.
  \item[Trajectory.] Line chart of $\hat{\theta}$ across the session,
    with hover tooltips showing the question ID and KC at each data point.
    The chart makes the adaptive engine's continuous ability estimation
    visible to the learner.
  \item[Skills.] Per-KC mastery bar chart showing $p(\text{Learned})$ for
    all 13 KCs, colour-coded by mastery status (mastered / in progress /
    not started).
  \item[Misconceptions.] Distractor frequency table showing the most common
    wrong-answer attractors per KC, enabling learners to identify specific
    knowledge gaps for targeted revision.
  \item[Share.] A shareable performance card (PNG export) for the leaderboard
    opt-in flow.
\end{description}

\subsection{Leaderboard}
\label{subsec:impl_leaderboard}

The leaderboard ranks learners by total XP, filtered to users who have
explicitly opted in via the \texttt{User.leaderboardOptIn} consent flag.
Consent is confirmed via a server-side double-check on every leaderboard
read: learners who withdraw consent are immediately removed from the ranked
display. Rank changes between sessions are animated using CSS transition
classes, providing visual feedback on relative performance improvement.

% ─────────────────────────────────────────────────────────────────────────────
\section{Challenges Encountered and Deviations from Plan}
\label{sec:impl_challenges}
% ─────────────────────────────────────────────────────────────────────────────

\subsection{EAP Grid Resolution}
\label{subsec:impl_eap_resolution}

The methodology chapter (Section~\ref{subsec:meth_eap}) specified an EAP
grid of 81 points with step $0.1$. Self-pilot testing revealed that this
resolution produced visible ``staircase'' discontinuities in the
$\hat{\theta}$ trajectory for high-ability learners, where the posterior
peak jumped between adjacent grid points rather than tracking smoothly. The
grid was refined to 161 points with step $0.05$, which resolved the artefact.
The computational cost increase (approximately $2\times$) was acceptable
given that EAP runs on the server per-question and not in the browser.

\subsection{Review Material Bug}
\label{subsec:impl_review_bug}

The ``Review Material'' feature (instruction mode between KC mastery and the
next KC introduction) triggers an unintended \texttt{setAppState('start')}
call in certain render paths within \texttt{InstructionMode.tsx}, causing
the session to reset to the start screen. This bug was identified during
usability testing and is documented as a known issue. It does not affect the
adaptive engine's core question selection or BKT update logic; it is confined
to the post-mastery instructional interstitial. A fix was scoped for Sprint 9
but deprioritised in favour of evaluation data collection. The bug is
disclosed here as an honest account of delivered system quality.

\subsection{K-Means Cold-Start: Evaluated and Superseded}
\label{subsec:impl_kmeans}

The interim report described a K-means clustering approach to cold-start
initialisation, in which new learners would be assigned to a cluster based
on demographic features and initialised with the cluster's median $\hat{\theta}$.
Following implementation of the pre-test seeding mechanism
(Section~\ref{sec:meth_pretest}), K-means was evaluated and superseded.
Pre-test seeding provides direct per-KC ability evidence rather than a
demographic proxy, is more interpretable, and avoids the ethical concern of
conditioning initial learning experience on demographic cluster assignment.
The K-means approach is therefore absent from the final implementation;
this section documents the deviation explicitly.

\subsection{Prisma React 18 Strict Mode Double-Invocation}
\label{subsec:impl_strict_mode}

React 18 Strict Mode double-invokes component render functions and
\texttt{useEffect} hooks in the development environment to surface
side-effect bugs. This caused apparent duplicate API calls in development
logs, which were initially misdiagnosed as a session state bug. Confirmed
to be a development-only behaviour with no impact on production sessions;
documented here to prevent re-investigation.

% ─────────────────────────────────────────────────────────────────────────────
\section{Requirements Traceability}
\label{sec:impl_traceability}
% ─────────────────────────────────────────────────────────────────────────────

\begin{table}[H]
\centering
\caption{Requirements traceability matrix: functional requirements to
implementation artefacts.}
\label{tab:traceability}
\renewcommand{\arraystretch}{1.2}
\begin{tabularx}{\textwidth}{lXlX}
\toprule
\textbf{Req.} & \textbf{Requirement} & \textbf{Status} & \textbf{Artefact} \\
\midrule
FR01 & Adaptive question selection using IRT 3PL & Done &
  \texttt{irt.ts:selectNextQuestion} \\
FR02 & Per-KC knowledge state tracking (BKT) & Done &
  \texttt{bkt.ts:updateBKT} \\
FR03 & Bloom-gated question escalation & Done &
  \texttt{next-question/route.ts} \\
FR04 & Pre-test ability seeding & Done &
  \texttt{irt.ts:estimateThetaFromAssessment} \\
FR05 & Cross-session variant deduplication & Done &
  \texttt{UserVariantSeen} model \\
FR06 & Graduated hint system with credit penalty & Done &
  \texttt{hints/route.ts}; \texttt{bkt.ts:updateBKT} \\
FR07 & BKT decay detection + successive relearning & Done &
  \texttt{next-question/route.ts} (decay boost) \\
FR08 & Session persistence and resume & Done &
  \texttt{Session.kcStates} JSON field \\
FR09 & SUS usability instrument & Done &
  Post-session questionnaire \\
FR10 & Leaderboard with consent opt-in & Done &
  \texttt{leaderboard/route.ts}; \texttt{User.leaderboardOptIn} \\
FR11 & NIST 800-63B compliant authentication & Done &
  Supabase Auth + bcrypt + rate limiting \\
FR12 & Analytics event log & Done &
  \texttt{AnalyticsEvent} model \\
FR13 & 7-day retention assessment & Partial &
  \texttt{Assessment.type = 'delayed\_post\_test'} (infrastructure
  implemented; data collection limited by evaluation timeline) \\
FR14 & Multi-domain KC support & Deferred &
  Architecture supports; content not authored \\
\bottomrule
\end{tabularx}
\end{table}

% ─────────────────────────────────────────────────────────────────────────────
\section{Deployment and Continuous Integration}
\label{sec:impl_ci}
% ─────────────────────────────────────────────────────────────────────────────

The build pipeline executes the following steps on every push to the main
branch:

\begin{enumerate}[leftmargin=2em]
  \item \texttt{prisma generate} — regenerates the Prisma client from the
        current schema, ensuring type safety is up to date.
  \item \texttt{tsc --noEmit} — TypeScript strict-mode type check. Build
        fails on any type error.
  \item \texttt{next build} — Next.js production build. Fails on any build
        error or unresolved import.
  \item Vercel deployment — if all previous steps pass, Vercel deploys the
        new build to the preview URL (for feature branches) or the production
        URL (for merges to main).
\end{enumerate}

Schema migrations are applied manually using \texttt{prisma db push} (for
rapid iteration on the development database) or \texttt{prisma migrate deploy}
(for production-safe migration with full rollback support). This two-track
migration strategy was an intentional trade-off: the development speed of
\texttt{db push} was prioritised during early sprints, with the switch to
formal migration files coinciding with the Sprint 7 security hardening phase.

% ─────────────────────────────────────────────────────────────────────────────
\section{Testing Strategy}
\label{sec:impl_testing}
% ─────────────────────────────────────────────────────────────────────────────

Three layers of testing were implemented to validate system correctness across
algorithmic, integration, and regression dimensions.

\begin{description}
  \item[Unit tests.] The IRT and BKT algorithm modules (\texttt{irt.ts},
    \texttt{bkt.ts}) are covered by 79 unit tests, validating: the 3PL
    probability function at boundary $\hat{\theta}$ values; EAP convergence
    on synthetic correct/incorrect response sequences; BKT posterior
    monotonicity under correct and incorrect response streams; credit-factor
    interpolation at all three hint levels ($cf \in \{1.0, 0.85, 0.65\}$);
    and Bloom gate threshold enforcement. Tests are authored in Jest and
    execute as part of the local development workflow. The algorithm modules'
    pure-function architecture --- no database calls, deterministic given the
    same inputs --- makes them directly amenable to unit testing without mocking.

  \item[Integration testing.] The end-to-end session flow was validated
    through a structured developer self-pilot on 2026-02-26: a 13-question
    pre-test followed by a full adaptive session, verifying that the EAP
    estimator, BKT update pipeline, item selector, and session state
    persistence all behaved correctly under realistic response patterns.
    This session also provided the empirical prior parameters
    ($\hat{\theta}_0 = -0.780$, $\sigma_0 = 0.543$) used by the live system.

  \item[Regression testing via CI.] TypeScript strict-mode type checking
    (\texttt{tsc --noEmit}) and the Next.js production build execute
    automatically on every push to \texttt{main}. The build fails on any
    type error or unresolved import, providing automated regression detection
    across all API routes and components. Feature branches are deployed to
    isolated Vercel preview URLs before merge, enabling manual integration
    testing without affecting the production environment.
\end{description}

The primary quality assurance gap is the absence of automated end-to-end
browser tests (Playwright or Cypress). UI regression testing relied on
manual verification against Vercel preview deployments. This limitation is
disclosed per the CSSE quality assurance principle: honest reporting of
defects and gaps, not a claim of defect-free delivery.

% ─────────────────────────────────────────────────────────────────────────────
\section{Accessibility Implementation}
\label{sec:impl_accessibility}
% ─────────────────────────────────────────────────────────────────────────────

The interface was developed to meet WCAG 2.1 Level AA accessibility guidelines
\parencite{w3c2018}. The following controls were implemented:

\begin{description}
  \item[ARIA labels.] All icon-only interactive elements --- the hint panel
    toggle, pause button, tab navigation icons, and progress indicators ---
    carry \texttt{aria-label} attributes describing their function, ensuring
    screen reader compatibility for learners using assistive technology.
  \item[Colour contrast.] Foreground text on authentication forms and answer
    option buttons was reviewed to meet the WCAG 4.5:1 minimum contrast ratio
    for normal text. Answer chips use a dark background with high-contrast
    light text, maintaining legibility under the system's dark theme across
    all feedback states (neutral, correct, incorrect).
  \item[Keyboard navigation.] Answer options, hint panels, and navigation
    controls are rendered as native \texttt{<button>} elements, accessible
    via Tab and activated via Enter or Space, preserving browser-native
    keyboard interaction without custom keyboard event handlers.
  \item[Reduced motion.] Framer Motion animation components (rank-change
    animation on the leaderboard) respect the operating system's
    \texttt{prefers-reduced-motion} media query; transitions are suppressed
    when the user has enabled the reduce-motion accessibility preference.
\end{description}

A full automated WCAG 2.1 AA audit (e.g., via axe-core or Lighthouse
accessibility scoring) was not conducted within the project timeline;
accessibility improvements were applied through manual review and iterative
correction. A comprehensive audit is identified as a prerequisite for any
production deployment targeting secondary school learners, where the public
sector bodies accessibility regulations may apply.

% ─────────────────────────────────────────────────────────────────────────────
\section{Self-Pilot Calibration}
\label{sec:impl_selfpilot}
% ─────────────────────────────────────────────────────────────────────────────

Before external participant recruitment, the developer conducted a full
self-pilot session on 2026-02-26. This served two purposes: (a) functional
verification of the end-to-end session flow under realistic conditions, and
(b) empirical derivation of the novice prior parameters
($\hat{\theta}_0 = -0.780$, $\sigma_0 = 0.543$). The resulting EAP estimate
after a 13-question pre-test was adopted as the system's default prior,
replacing the arbitrary $\mathcal{N}(0, 1)$ prior used during development.
This approach follows the recommendation of \textcite{baker2004} that EAP
priors should be empirically grounded where possible, even from small pilots,
rather than assigned on theoretical grounds alone.
