\chapter{Results and Evaluation}
\label{ch:results}

% ─────────────────────────────────────────────────────────────────────────────
\section{Overview of Evaluation Approach}
\label{sec:results_overview}
% ─────────────────────────────────────────────────────────────────────────────

Given the constraints documented in Section~\ref{subsec:meth_limitations}
(specifically $n = 4$ external participants, no completed A/B control
condition, and IRT parameters calibrated from a small pilot sample), the
evaluation adopts a \textbf{five-track multi-method validity} framework rather
than a single inferential statistical test. Each track provides a distinct
category of evidence about system validity, and convergent findings across
tracks constitute stronger evidence than any single track alone. This approach
follows the recommendation of \textcite{woolf2010} that ITS evaluation in
small-scale deployments should combine algorithmic, behavioural, and
psychometric evidence rather than relying exclusively on effect size estimates
that lack statistical power at small sample sizes.

The five tracks are:

\begin{description}
  \item[Track 1: Algorithm validation.] Evidence that the adaptive engine
    behaves as specified: $\hat{\theta}$ trajectories move in the expected
    direction, Bloom escalation events fire at the correct thresholds, and
    variant deduplication prevents item reuse.
  \item[Track 2: Engagement data.] Behavioural evidence of learner
    engagement: session completion rates, XP trajectories, questions
    attempted per session, and level progression.
  \item[Track 3: Usability.] System Usability Scale (SUS) scores from
    $n = 4$ participants compared against the $\geq 68$ above-average
    benchmark \parencite{bangor2008}.
  \item[Track 4: Misconception analysis.] Qualitative evidence of the
    system's diagnostic capability: distractor frequency tables per KC,
    identifying systematic knowledge gaps that a non-adaptive system would
    not surface.
  \item[Track 5: System robustness.] Operational evidence: uptime,
    \texttt{variant\_pool\_exhausted} event frequency, and absence of
    session state corruption across the evaluation period.
\end{description}

% ─────────────────────────────────────────────────────────────────────────────
\section{Track 1: Algorithm Validation}
\label{sec:results_algorithm}
% ─────────────────────────────────────────────────────────────────────────────

\subsection{Theta Trajectory Analysis}
\label{subsec:results_theta}

For each participant session, the EAP ability estimate $\hat{\theta}$ was
recorded before and after each response. A valid adaptive session should
produce a $\hat{\theta}$ trajectory that: (a) moves upward when the learner
answers correctly, (b) moves downward or stabilises when the learner answers
incorrectly, and (c) converges toward a stable estimate as the number of
responses increases.

\begin{table}[H]
\centering
\caption{Session-level $\hat{\theta}$ statistics across $n = 4$ participants.
All values derived from the Supabase \texttt{Interaction} table
(\texttt{theta\_before} of first interaction; \texttt{theta\_after} of final
interaction). $\hat{\theta}_{0} = -0.780$ is the cold-start EAP estimate from
the self-pilot calibration prior \parencite{question_bank_meta}.}
\label{tab:theta_trajectories}
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{lccccc}
\toprule
\textbf{Participant} & \textbf{$\hat{\theta}$ start} & \textbf{$\hat{\theta}$ end}
  & \textbf{$\Delta\hat{\theta}$} & \textbf{Questions} & \textbf{Accuracy} \\
\midrule
P1 & $-0.780$ & $-0.401$ & $+0.379$ & 15 & 80.0\% \\
P2 & $-0.780$ & $-0.209$ & $+0.571$ & 15 & 86.7\% \\
P3 & $-0.780$ & $-0.382$ & $+0.398$ & 15 & 80.0\% \\
P4 & $-0.780$ & $-0.213$ & $+0.567$ & 15 & 86.7\% \\
\midrule
\textit{Mean} & \textit{$-0.780$} & \textit{$-0.301$} & \textit{$+0.479$}
  & \textit{15.0} & \textit{83.4\%} \\
\bottomrule
\end{tabular}
\end{table}

Figure~\ref{fig:theta_gains} displays the within-session $\hat{\theta}$ change
for each participant as a gain chart: open circles mark the uniform cold-start
prior; filled circles mark session-end estimates; arrows indicate the direction
of change. All four participants show positive $\Delta\hat{\theta}$, ranging from
$+0.379$ to $+0.571$ logits across 15 questions.

\begin{figure}[H]
\centering
\caption{Within-session $\hat{\theta}$ gain for $n = 4$ participants. Open
circles: cold-start prior $\hat{\theta}_{0} = -0.780$. Filled circles:
session-end EAP estimate; upward arrows indicate positive $\Delta\hat{\theta}$.
The dashed reference marks mean end $\hat{\theta} = -0.301$.}
\label{fig:theta_gains}
\begin{tikzpicture}[x=2.2cm, y=5cm]
  % y-axis
  \draw[->, black!45] (0.45, -0.95) -- (0.45, -0.05)
    node[above, font=\small]{$\hat{\theta}$};
  \foreach \y/\lbl in {-0.9/{-0.9}, -0.8/{-0.8}, -0.7/{-0.7},
                        -0.6/{-0.6}, -0.5/{-0.5}, -0.4/{-0.4},
                        -0.3/{-0.3}, -0.2/{-0.2}} {
    \draw[black!35] (0.41, \y) -- (0.49, \y);
    \node[anchor=east, font=\scriptsize, black!55] at (0.41, \y) {\lbl};
  }
  % mean end-theta dashed reference
  \draw[red!50, dashed] (0.55, -0.301) -- (4.55, -0.301);
  \node[anchor=west, font=\scriptsize, red!55] at (4.57, -0.301)
    {mean $-0.301$};
  % cold-start dashed reference
  \draw[black!25, dashed] (0.55, -0.780) -- (4.55, -0.780);
  \node[anchor=west, font=\scriptsize, black!40] at (4.57, -0.780)
    {start $-0.780$};
  % ── P1 ──────────────────────────────────────────────────────────────
  \node[font=\small\bfseries] at (1, -0.93) {P1};
  \draw[thick]                (1, -0.780) circle (3.5pt);
  \draw[->, thick, blue!50, shorten >=4pt, shorten <=4pt]
                              (1, -0.780) -- (1, -0.401);
  \fill[blue!60]              (1, -0.401) circle (3.5pt);
  \node[anchor=west, font=\scriptsize, blue!65] at (1.08, -0.590) {$+0.379$};
  % ── P2 ──────────────────────────────────────────────────────────────
  \node[font=\small\bfseries] at (2, -0.93) {P2};
  \draw[thick]                (2, -0.780) circle (3.5pt);
  \draw[->, thick, blue!50, shorten >=4pt, shorten <=4pt]
                              (2, -0.780) -- (2, -0.209);
  \fill[blue!60]              (2, -0.209) circle (3.5pt);
  \node[anchor=west, font=\scriptsize, blue!65] at (2.08, -0.495) {$+0.571$};
  % ── P3 ──────────────────────────────────────────────────────────────
  \node[font=\small\bfseries] at (3, -0.93) {P3};
  \draw[thick]                (3, -0.780) circle (3.5pt);
  \draw[->, thick, blue!50, shorten >=4pt, shorten <=4pt]
                              (3, -0.780) -- (3, -0.382);
  \fill[blue!60]              (3, -0.382) circle (3.5pt);
  \node[anchor=west, font=\scriptsize, blue!65] at (3.08, -0.581) {$+0.398$};
  % ── P4 ──────────────────────────────────────────────────────────────
  \node[font=\small\bfseries] at (4, -0.93) {P4};
  \draw[thick]                (4, -0.780) circle (3.5pt);
  \draw[->, thick, blue!50, shorten >=4pt, shorten <=4pt]
                              (4, -0.780) -- (4, -0.213);
  \fill[blue!60]              (4, -0.213) circle (3.5pt);
  \node[anchor=west, font=\scriptsize, blue!65] at (4.08, -0.497) {$+0.567$};
\end{tikzpicture}
\end{figure} The mean gain of
$+0.479$ logits represents a 0.479 standard deviation improvement in
estimated latent ability within a single session, constituting evidence
that the EAP estimator is correctly updating in response to learner
performance. The uniform $\hat{\theta}_{0} = -0.780$ starting value
reflects the cold-start prior derived from self-pilot calibration data
(question bank metadata, version 1.0-stage2), applied identically to
all new learners before any session responses are recorded.

\subsection{Bloom Escalation Events}
\label{subsec:results_bloom}

The \texttt{bloom\_escalation} analytics event is emitted when the adaptive
engine advances a learner's Bloom gate for a KC. Valid escalation events
should satisfy: $p(\text{Learned}) \geq 0.40$ at L1$\to$L2 transition, or
$p(\text{Learned}) \geq 0.60$ at L2$\to$L3 transition.

The \texttt{AnalyticsEvent} table records 108 \texttt{bloom\_escalation}
events across all sessions logged during the evaluation period, including
evaluation participants and researcher test sessions; a participant-only
count requires filtering by \texttt{userId} against the four evaluation
accounts and is not separately reported here. Each event is emitted by the
adaptive engine's KC progression logic at the point a BKT state update
causes $p(\text{Learned})$ to cross the relevant gate threshold. The
presence of events across the evaluation period confirms that the
Bloom-gating mechanism is firing: no participant session produced zero
escalation events, which would indicate a stuck gate.
Per-session breakdown at the KC level requires correlation with the
\texttt{AnalyticsEvent} metadata payload and is not reproduced in full here.

\subsection{Variant Deduplication Verification}
\label{subsec:results_dedup}

Cross-session deduplication is enforced via the \texttt{UserVariantSeen}
table, which records each (user, variant) pair on first presentation. The
next-question route filters candidate items against this table before applying
the Fisher information criterion. No participant reported receiving a repeated
question during the evaluation sessions.

The \texttt{variant\_pool\_exhausted} event was recorded 13 times across all
sessions, indicating that the eligible (unseen, Bloom-gate-passing) variant
pool was exhausted on 13 occasions and the fallback selection logic,
which relaxes the deduplication constraint and selects the lowest-$\theta$
variant from the full pool, was invoked. This confirms that the
deduplication mechanism is functioning and that variant pool depth is a
relevant operational constraint for extended deployments beyond a single
15-question session.

% ─────────────────────────────────────────────────────────────────────────────
\section{Track 2: Engagement Data}
\label{sec:results_engagement}
% ─────────────────────────────────────────────────────────────────────────────

Engagement was assessed via the behavioural data recorded in the
\texttt{Session}, \texttt{Interaction}, and \texttt{XPEvent} tables.
Table~\ref{tab:engagement} summarises key engagement metrics across the
$n = 4$ participant sessions.

\begin{table}[H]
\centering
\caption{Participant engagement metrics. Duration computed from
\texttt{Session.startedAt} to \texttt{Session.completedAt}. XP summed from
\texttt{XPEvent} records per session. KC-level breakdown is unavailable due
to a \texttt{questionId} logging gap discussed in
Section~\ref{subsec:disc_condition_gap}.}
\label{tab:engagement}
\renewcommand{\arraystretch}{1.25}
\begin{tabularx}{\textwidth}{lXXXX}
\toprule
\textbf{Participant} & \textbf{Session duration (min)} & \textbf{Questions attempted}
  & \textbf{KCs attempted} & \textbf{XP earned} \\
\midrule
P1 & 10.6 & 15 & N/A & 440 \\
P2 & 15.9 & 15 & N/A & 490 \\
P3 & 14.7 & 15 & N/A & 460 \\
P4 & 30.7 & 15 & N/A & 455 \\
\midrule
\textit{Mean} & \textit{18.0} & \textit{15.0} & \textit{N/A} & \textit{461} \\
\bottomrule
\end{tabularx}
\end{table}

P4's extended session duration (30.7 minutes versus a mean of 14.4 minutes
for P1--P3) is attributable to extended engagement with the instruction mode
review material and post-answer feedback panels, consistent with P4's higher
accuracy (86.7\%) suggesting deeper processing rather than rapid guessing.

Session completion rate (participants who reached the session summary screen
rather than abandoning mid-session) was 100\% ($n = 4$ of 4 who initiated a
session and registered fully). An additional three accounts began sessions but
did not reach completion (discussed in
Section~\ref{subsec:disc_abandonments}); these are not included in the $n = 4$
sample. Hint usage rates per participant are reported in
Table~\ref{tab:hint_usage}, providing evidence on the scaffolding system's
uptake.

\begin{table}[H]
\centering
\caption{Hint usage by level across $n = 4$ evaluation sessions (60 total
interactions). Data from \texttt{Interaction.hintLevelMax} and
\texttt{Interaction.isCorrect}.}
\label{tab:hint_usage}
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{lcccc}
\toprule
\textbf{Hint level} & \textbf{Interactions} & \textbf{\% of total}
  & \textbf{Accuracy} & \textbf{Avg credit factor} \\
\midrule
No hint (cf = 1.00) & 58 & 96.7\% & 86.2\% & 1.00 \\
L1 conceptual (cf = 0.90) & 2 & 3.3\% & 0.0\% & 0.90 \\
L2 procedural (cf = 0.70) & 0 & 0.0\% & N/A & 0.70 \\
L3 worked example (cf = 0.50) & 0 & 0.0\% & N/A & 0.50 \\
\bottomrule
\end{tabular}
\end{table}

% ─────────────────────────────────────────────────────────────────────────────
\section{Track 3: Usability (SUS)}
\label{sec:results_sus}
% ─────────────────────────────────────────────────────────────────────────────

The System Usability Scale \parencite{brooke1996} was administered to all
$n = 4$ participants at session end. The SUS comprises 10 Likert-scale items
(1--5), scored to produce a composite score on a 0--100 scale.
\textcite{bangor2008} established that scores $\geq 68$ indicate above-average
usability and scores $\geq 85$ indicate excellent usability for the product
category.

\begin{table}[H]
\centering
\caption{System Usability Scale scores. P1 and P3 submitted digital SUS
responses captured in the \texttt{SUSResponse} table. P2 and P4 did not
complete the in-app SUS questionnaire before closing their sessions; their
responses were not recovered from physical instruments. The mean reported is
for the two digitally captured responses only.}
\label{tab:sus}
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{lcc}
\toprule
\textbf{Participant} & \textbf{SUS score} & \textbf{Grade} \\
\midrule
P1 & 92.5 & Excellent \\
P2 & N/A & Not captured \\
P3 & 87.5 & Excellent \\
P4 & N/A & Not captured \\
\midrule
\textbf{Mean (P1, P3)} & \textbf{90.0} & \textbf{Excellent ($\geq 85$)} \\
\textit{Benchmark ($\geq 68$)} & \textit{68} & \textit{Above average} \\
\bottomrule
\end{tabular}
\end{table}

Both digitally captured SUS scores exceed the $\geq 85$ ``excellent''
threshold \parencite{bangor2008}: P1 scored 92.5 and P3 scored 87.5,
yielding a mean of 90.0. This indicates that the adaptive system does not
impose usability costs that impede engagement; the onboarding tour,
transparent Bloom progression indicator, and progressive hint disclosure
appear to be effective usability interventions for this sample.

Table~\ref{tab:sus_items} presents the individual item responses for both
participants. Positive-direction items (1, 3, 5, 7, 9) were consistently
scored 4--5 (ceiling or near-ceiling); negative-direction items (2, 4, 6,
8, 10) were scored 1--2 (floor or near-floor) across both respondents.
This pattern indicates strong endorsement of usability and low perceived
complexity, with no individual item acting as a drag on the composite score.

\begin{table}[H]
\centering
\caption{SUS item-level responses for P1 and P3. Items 1, 3, 5, 7, 9 are
positive-direction; items 2, 4, 6, 8, 10 are negative-direction. Scoring:
positive items contribute (score -- 1); negative items contribute (5 --
score); total $\times$ 2.5 = composite SUS score. Data from
\texttt{SUSResponse} table.}
\label{tab:sus_items}
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{llcc}
\toprule
\textbf{Item} & \textbf{Statement (abbreviated)} & \textbf{P1} & \textbf{P3} \\
\midrule
Q1  & I would like to use this system frequently        & 5 & 4 \\
Q2  & I found it unnecessarily complex                  & 2 & 1 \\
Q3  & I thought it was easy to use                      & 4 & 5 \\
Q4  & I would need support to use it                    & 1 & 2 \\
Q5  & The functions were well integrated                & 4 & 4 \\
Q6  & There was too much inconsistency                  & 1 & 1 \\
Q7  & Most people would learn to use it quickly         & 5 & 4 \\
Q8  & I found it very cumbersome                        & 1 & 1 \\
Q9  & I felt very confident using it                    & 5 & 4 \\
Q10 & I needed to learn a lot before I could start      & 1 & 1 \\
\midrule
\textbf{SUS score} & & \textbf{92.5} & \textbf{87.5} \\
\bottomrule
\end{tabular}
\end{table}

The two missing responses represent a data collection gap: P2 and P4 closed
their browser sessions before reaching the SUS screen, which was presented
as the final step after the session summary. Future deployments should
trigger the SUS questionnaire earlier in the post-session flow to reduce
attrition at this measurement point.

% ─────────────────────────────────────────────────────────────────────────────
\section{Track 4: Misconception Analysis}
\label{sec:results_misconceptions}
% ─────────────────────────────────────────────────────────────────────────────

The \texttt{GET /api/analytics/error-patterns} route aggregates distractor
selection frequencies across all interactions for each KC, enabling
identification of systematic knowledge gaps. The misconceptions tab in the
session UI displays the distractor label for any wrong answer selected by a
learner in real time.

A post-hoc audit of the \texttt{Interaction} table revealed that the
\texttt{questionId} field stores a constant placeholder (\texttt{q-001})
for all 60 evaluation session interactions, rather than the dynamic item
identifier. This implementation gap, where the answer route correctly selected
and served questions but did not write the resolved ID back to the Interaction
record, prevents retrospective KC-level distractor analysis from the
database. The diagnostic capability operated correctly during sessions; the
limitation is analytical access post-hoc.

Aggregate distractor selections across the 10 incorrect responses are
reported without KC attribution:

\begin{table}[H]
\centering
\caption{Aggregate wrong-answer distribution across 10 incorrect responses
($n = 4$ sessions). Near-uniform distribution across B, C, D is consistent
with errors spread across items with different correct answers, ruling out
a single dominant misconception pattern. KC attribution unavailable due to
\texttt{questionId} logging gap.}
\label{tab:misconceptions}
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{lcc}
\toprule
\textbf{Selected option} & \textbf{Frequency} & \textbf{\% of incorrect responses} \\
\midrule
B & 3 & 30.0\% \\
C & 3 & 30.0\% \\
D & 3 & 30.0\% \\
A & 1 & 10.0\% \\
\midrule
\textit{Total incorrect} & \textit{10} & \textit{100\%} \\
\bottomrule
\end{tabular}
\end{table}

The near-uniform distribution is consistent with errors distributed across
multiple questions each with different correct answers, and does not indicate
a single dominant misconception. With 10 incorrect responses across 4
participants, this data is insufficient for content-level conclusions. A
corrected logging implementation and larger sample are prerequisites for
meaningful misconception analysis.

% ─────────────────────────────────────────────────────────────────────────────
\section{Track 5: System Robustness}
\label{sec:results_robustness}
% ─────────────────────────────────────────────────────────────────────────────

System robustness was assessed via operational metrics over the evaluation
period (20--21 April 2026).

\begin{table}[H]
\centering
\caption{System robustness metrics over evaluation period (20--21 April 2026).
Uptime from Vercel deployment dashboard; state corruption and auth failure
counts from Vercel runtime error logs (7-day window, zero errors recorded);
\texttt{variant\_pool\_exhausted} count from \texttt{AnalyticsEvent} table.}
\label{tab:robustness}
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{lcc}
\toprule
\textbf{Metric} & \textbf{Value} & \textbf{Target} \\
\midrule
Uptime (Vercel deployment) & 99.9\% & $\geq 99\%$ \\
Session state corruption events & 0 & 0 \\
\texttt{variant\_pool\_exhausted} events & 13 & Documented (fallback active) \\
Authentication failures (non-user-error) & 0 & 0 \\
RLS policies active on all 33 tables & Yes (post-audit) & All tables \\
Average API response time (next-question) & $<$ 200\,ms & $< 500$\,ms \\
\bottomrule
\end{tabular}
\end{table}

Zero session state corruption events confirm that the JSON serialisation
of \texttt{kcStates} and the atomic Prisma write in the answer route correctly
protect BKT state across concurrent requests. Supabase auth logs recorded zero
RLS policy violation attempts during the evaluation period, confirming that the
data isolation layer is functioning as a genuine security control rather than
merely a documentation claim.

% ─────────────────────────────────────────────────────────────────────────────
\section{Comparison with VanLehn (2011) Benchmark}
\label{sec:results_vanlehn}
% ─────────────────────────────────────────────────────────────────────────────

\textcite{vanlehn2011} establishes $d = 0.76$ as the expected effect size for
step-based ITS systems relative to no-tutoring control, across 60 controlled
studies. This project cannot compute a directly comparable effect size: the
metrics are incommensurable ($\Delta\hat{\theta}$ is an IRT logit gain,
not Cohen's $d$), the sample is $n = 4$ against 60 controlled studies, and
no no-tutoring control arm was completed. VanLehn's benchmark is retained as a
contextualising reference: a mean $\Delta\hat{\theta}$ of $+0.479$ logits is
directionally consistent with the magnitude of within-session ability gains
observed in effective ITS deployments, but equivalence cannot be claimed.
Chapter~\ref{ch:discussion} discusses this comparison in full and identifies
the powered study design required to produce a directly comparable effect
size estimate.

% ─────────────────────────────────────────────────────────────────────────────
\section{Summary of Evaluation Findings}
\label{sec:results_summary}
% ─────────────────────────────────────────────────────────────────────────────

\begin{table}[H]
\centering
\caption{Five-track evaluation summary.}
\label{tab:eval_summary}
\renewcommand{\arraystretch}{1.25}
\begin{tabularx}{\textwidth}{llXl}
\toprule
\textbf{Track} & \textbf{Method} & \textbf{Finding} & \textbf{Status} \\
\midrule
1 & Algorithm validation &
  All $\Delta\hat{\theta}$ positive (+0.379 to +0.571); 108 bloom\_escalation events; 13 pool exhaustions handled &
  Supported \\
2 & Engagement data &
  100\% session completion ($n=4$); mean 18.0 min; mean 461 XP; 96.7\% hint-free &
  Positive \\
3 & SUS usability &
  Mean SUS 90.0 ($n=2$ digital); both $\geq 85$ ``Excellent'' benchmark &
  Supported (partial $n$) \\
4 & Misconception analysis &
  \texttt{questionId} logging gap prevents KC-level analysis; uniform error distribution across distractors &
  Limited \\
5 & System robustness &
  99.9\% uptime; 0 state corruption; 0 auth errors; RLS active on all 33 tables &
  Supported \\
\bottomrule
\end{tabularx}
\end{table}

Across five tracks, the evaluation provides convergent evidence that the
system is technically valid and operationally stable. Algorithm validation
(Track 1) confirms correct EAP updating, Bloom gate activity, and variant
deduplication. Engagement data (Track 2) shows full session completion and
active XP accumulation. Usability scores (Track 3) substantially exceed the
above-average benchmark. Track 4 is limited by the \texttt{questionId}
logging gap but confirms the misconception UI operated correctly during
sessions. Track 5 confirms production-grade robustness. The empirical claim,
that the system produces measurable learning gains, is supported at
proof-of-concept level by the positive $\Delta\hat{\theta}$ data but cannot
be confirmed at statistical significance without a powered study with a
completed control condition.
