\chapter{Discussion}
\label{ch:discussion}

% ─────────────────────────────────────────────────────────────────────────────
\section{Does the System Answer the Research Question?}
\label{sec:disc_rq}
% ─────────────────────────────────────────────────────────────────────────────

The central research question of this project is: \textit{can a hybrid IRT
3PL + BKT adaptive engine, implemented in a production web application,
deliver personalised question sequencing that produces measurable learning
gains within a single session for UK KS3/GCSE Geography content?}

The system as implemented provides a technically complete answer to this
question. The adaptive engine correctly selects items based on Fisher
information at the learner's current $\hat{\theta}$, enforces Bloom-level
cognitive prerequisites, tracks per-KC knowledge state via BKT, and applies
the credit factor modification for hint-assisted responses. All components
behave as specified, as confirmed by the algorithm validation evidence in
Track 1 (Section~\ref{sec:results_algorithm}).

The empirical answer --- whether measurable learning gains actually occur --- is
provisionally supported by the within-session $\Delta\hat{\theta}$ data
collected from $n = 4$ participants, but cannot be stated with statistical
confidence given the sample size. The honest position is: the system is
\textit{technically validated} (it behaves correctly) and \textit{empirically
indicative} (preliminary data is directionally consistent with the expected
effect), but \textit{not yet inferentially confirmed} (a study with $n \geq
15$ and a control condition is required to produce a publishable effect size
estimate). This distinction between technical and empirical validation is
important and is frequently elided in student project reports; it is made
explicit here.

% ─────────────────────────────────────────────────────────────────────────────
\section{Comparison with Related Work}
\label{sec:disc_comparison}
% ─────────────────────────────────────────────────────────────────────────────

\subsection{VanLehn (2011)}
\label{subsec:disc_vanlehn}

\textcite{vanlehn2011} identifies step-based feedback as the critical
determinant of ITS effectiveness, with step-based systems achieving $d = 0.76$
versus $d = 0.31$ for answer-based systems. GeoMentor provides question-level
feedback (``correct / incorrect, here is the explanation'') rather than
step-level feedback (``you went wrong at this specific reasoning step'').
Strictly, this classifies the system as an \textit{answer-based} tutoring
system under VanLehn's taxonomy, suggesting an expected effect size closer to
$d = 0.31$ than $d = 0.76$ if a controlled study were conducted.

However, two features of this system move it toward the step-based category.
First, the three-level hint system progressively scaffolds the solution
pathway before the learner commits an answer, providing something closer to
step-level guidance than a simple post-hoc explanation. Second, the
misconception tab identifies the specific wrong-answer attractor selected by
the learner, enabling targeted reflection on the specific reasoning error made.
Whether these features are sufficient to produce step-based effect sizes is an
empirical question that the current evaluation cannot resolve. It is an honest
limitation that VanLehn's $d = 0.76$ benchmark, while used as a reference
throughout this report, may be an optimistic comparator for this system's
interaction granularity.

\subsection{Kulik and Fletcher (2016)}
\label{subsec:disc_kulik}

\textcite{kulik2016} report a weighted mean effect size of $d = 0.66$ for
systems implementing continuous real-time adaptation --- the category into which
GeoMentor falls, since $\hat{\theta}$ and $p(\text{Learned})$ are updated after
every single response. Their comparison with discrete checkpoint systems ($d =
0.28$) supports the design decision to implement per-response EAP updating
rather than a batch-update architecture. This is one area where the
implementation is at the upper bound of what the literature predicts for
systems of this class.

% ─────────────────────────────────────────────────────────────────────────────
\section{Scope Reduction: Strength Not Weakness}
\label{sec:disc_scope}
% ─────────────────────────────────────────────────────────────────────────────

The decision to scope down from multi-domain to single-domain Geography, and
subsequently to validate a single core KC workflow, might appear to weaken the
project's claims. The opposite case is argued here.

\textcite{desmarais2012} note that the most common failure mode in adaptive
system research is shallow multi-domain validation, in which a system is
demonstrated to work across many domains with weak evidence in each. Deep
single-domain validation --- rigorous psychometric calibration, detailed
misconception analysis, and multi-track validity evidence within one domain
--- produces more defensible research contributions. The geographic focus also
enables a contribution that the multi-domain version could not have made:
the demonstration that ITS methods transfer to a declarative-knowledge
non-STEM domain is itself a research finding, addressing Gap 1 identified in
the literature review (Section~\ref{sec:lr_gaps}).

Furthermore, the architecture is domain-agnostic by design: the adaptive
engine's KC graph, IRT parameter store, and BKT state management are
parameterised by KC identifiers and do not contain any Geography-specific
logic. Extending the system to additional domains requires authoring new KC
graphs and question banks, but no algorithmic changes. The scope reduction
is therefore a validation decision, not an architectural constraint.

% ─────────────────────────────────────────────────────────────────────────────
\section{Limitations}
\label{sec:disc_limitations}
% ─────────────────────────────────────────────────────────────────────────────

\subsection{Sample Size}
\label{subsec:disc_n4}

The most significant limitation of this evaluation is the sample size of
$n = 4$ external participants. Paired $t$-test at $n = 4$ has approximately
19\% statistical power at $d = 0.76$ and $\alpha = 0.05$ --- effectively
indistinguishable from chance in terms of hypothesis testing. The evaluation's
multi-method approach was adopted precisely because inferential statistics are
not viable at this sample size. The findings reported in Chapter~\ref{ch:results}
are descriptive and indicative; they establish proof-of-concept rather than
empirical proof of effect.

The target of $n = 15$ was not achieved due to the compressed evaluation
timeline following the Sprint 5 scope reduction. Recruitment was initiated
via University of Hull participant pools and departmental email lists;
response rates were lower than anticipated. This is a common challenge in
student project research and is not unique to this evaluation, but it is
acknowledged as the primary factor limiting the strength of the empirical
evidence.

\subsection{Absence of a Completed Control Condition}
\label{subsec:disc_control}

The within-subjects quasi-experimental design specified in the methodology
required each participant to experience both an adaptive and a static
(linear) condition. The static question-sequencing branch is correctly
implemented in the system: when a session's \texttt{condition} field is set
to \texttt{static}, the next-question route selects items by ascending
difficulty parameter $b$ rather than by Fisher information, providing a
fixed-difficulty linear progression through the item pool. What failed was
the assignment mechanism: as described in Section~\ref{subsec:disc_condition_gap},
all evaluation participants received the adaptive condition because the
condition assignment was not wired through to the signup pipeline.
VanLehn's meta-analytic benchmark was therefore used as a surrogate
comparator rather than an internal control arm. This means that any
observed within-session $\Delta\hat{\theta}$ cannot be attributed
specifically to the adaptive mechanism: it could reflect general learning
from Geography content regardless of sequencing. A two-condition study is
required to isolate the adaptation effect.

\subsection{Condition Assignment Implementation Gap}
\label{subsec:disc_condition_gap}

A late-stage audit of the codebase against dissertation claims revealed that
the \texttt{studyCondition} field, added to the \texttt{User} model at
Sprint~8, was never populated by the signup API route. The Prisma schema
default (\texttt{"adaptive"}) therefore silently applied to every registered
user, and the \texttt{ConditionExplainer} component presented participants
with a binary choice between conditions rather than communicating a
pre-assigned condition --- introducing self-selection bias at the point of
evaluation. The bug was identified by comparing the dissertation's claim
(``participants are randomly assigned to adaptive or static at account
creation'') against the actual signup route implementation, and corrected
before final submission. However, by the time of discovery all four
evaluation participants had already completed their sessions under the
adaptive condition.

This represents a gap between specification and implementation that is
instructive from a software engineering perspective. The schema was correct;
the data flow was incomplete; and the error was invisible during feature
development because the adaptive condition also happened to be the
behaviorally correct path. The gap would have been caught earlier by an
integration test asserting that \texttt{studyCondition} is non-null
after user creation, or by a pre-evaluation audit of the live database.
Both are identified as process improvements for future iterations.

\subsection{IRT Parameters from Small Calibration Sample}
\label{subsec:disc_calibration}

IRT item parameters were estimated from a small pilot calibration sample.
The 3PL model's discrimination parameter $a$ is particularly sensitive to
sparse data: with fewer than 100 responses per item, the standard errors of
$a$ estimates are large, meaning that the Fisher information criterion may
rank items differently from their true information order. The constrained
calibration approach (fixing $c = 0.25$, monitoring RMSEA) mitigates but does
not eliminate this uncertainty. A recalibration using data from the evaluation
sessions would produce more stable parameters and is identified as an
immediate next step for any continued development.

\subsection{Single Session per Participant}
\label{subsec:disc_single_session}

The evaluation captured single-session data for all $n = 4$ participants.
The BKT cross-session decay detection mechanism --- one of the system's novel
contributions --- could therefore not be evaluated empirically. Similarly,
the 7-day delayed retention assessment was administered to zero
participants within the available data collection window, providing no
data for retention analysis. Multi-session
longitudinal data collection is the most important methodological gap for
future work.

\subsection{Partial Session Abandonments}
\label{subsec:disc_abandonments}

Database records show three additional participants who began evaluation
sessions but did not reach the session completion screen: one participant
answered four questions, one answered eight, and one opened the system
without answering any questions before exiting. These partial sessions are
informative in two respects.

First, they confirm that the system was reachable and functional: participants
were able to authenticate, start a session, and receive adaptive question
selections during the evaluation window. The question-and-answer loop,
the hint scaffolding, and the $\hat{\theta}$ update pipeline all operated
correctly for these users; the data simply does not include a completed
session pair from which $\Delta\hat{\theta}$ can be computed. Their partial
$\hat{\theta}$ trajectories are stored in the database and available for
future analysis.

Second, the dropout pattern is consistent with what the literature reports
for voluntary digital learning deployments. \textcite{koedinger2012} note
that self-directed abandonment prior to session completion is a normative
behaviour in low-stakes adaptive learning contexts, and does not necessarily
indicate interface failure. Nevertheless, with only four completed sessions
in the sample, even partial dropout substantially affects statistical power.
Identifying and eliminating interface friction points that cause early
abandonment is a priority for any subsequent recruitment effort.

\subsection{Convenience Sample Generalisability}
\label{subsec:disc_generalisability}

University of Hull students are not representative of the UK KS3/GCSE student
population that the system targets. University students have higher average
prior knowledge of Geography (particularly at the declarative-knowledge level
tested at Bloom Level 1), greater familiarity with computer-mediated learning
interfaces, and different motivational profiles than secondary school learners.
These differences mean that the evaluation's usability and engagement findings
may overestimate performance with the intended population.

% ─────────────────────────────────────────────────────────────────────────────
\section{Threats to Validity}
\label{sec:disc_threats}
% ─────────────────────────────────────────────────────────────────────────────

\begin{description}
  \item[Internal validity.] The primary threat to internal validity is the
    absence of a control condition: any observed learning gains could be
    attributed to content exposure alone, test practice effects, or
    regression to the mean rather than to the adaptive mechanism. The
    within-subjects design partially mitigates this by using each participant
    as their own control, but without a completed static arm this mitigation
    is incomplete.

  \item[Construct validity.] The EAP $\hat{\theta}$ is a latent construct
    that cannot be directly observed. Its validity as a measure of geography
    knowledge depends on the quality of the IRT calibration, which is limited
    by the small calibration sample. Construct validity would be strengthened
    by demonstrating convergent validity between $\hat{\theta}$ and an
    external geography knowledge measure (e.g., a standardised GCSE practice
    paper score).

  \item[Ecological validity.] The evaluation was conducted under researcher-
    administered conditions, potentially producing demand characteristics:
    participants may have exerted greater effort than they would in
    self-directed use. Self-directed deployment over a longer period would
    provide more ecologically valid engagement data.

  \item[Statistical conclusion validity.] With $n = 4$, statistical conclusion
    validity is severely limited. No inferential statistical claims are made
    in this report; all numerical findings are descriptive. The risk of
    Type II error (failing to detect a genuine effect) is high.
\end{description}

% ─────────────────────────────────────────────────────────────────────────────
\section{What Could Not Be Delivered and Why}
\label{sec:disc_not_delivered}
% ─────────────────────────────────────────────────────────────────────────────

Three commitments from the original project proposal were not delivered:

\textbf{Multi-domain support.} The original proposal described an adaptive
system covering STEM and Humanities domains. This was evaluated and superseded
by the scope reduction decision (Section~\ref{subsec:meth_scope_reduction}).
The decision was driven by: (a) the infeasibility of calibrating separate IRT
parameters for multiple domains within the available timeline; (b) the
dilution of evaluation statistical power across subgroups. The architecture
supports multi-domain extension; the content does not yet exist.

\textbf{Completed A/B randomised controlled trial.} The original evaluation
plan called for a completed no-adaptation control condition. The static
question-sequencing branch was implemented (items sorted by ascending
difficulty $b$ rather than Fisher information), but the condition
assignment pipeline was not fully wired at signup, meaning all evaluation
participants received the adaptive condition in practice
(Section~\ref{subsec:disc_condition_gap}). The comparison with the static
arm therefore relies on VanLehn's meta-analytic benchmark as a surrogate
rather than an internal control condition. A two-arm study with confirmed
random assignment, verified at the point of data collection via database
audit, is the most important methodological addition for any follow-up
investigation.

\textbf{$n = 15$ participant target.} Participant recruitment fell short of
the target sample size. Contributing factors include: the compressed timeline
between the scope reduction decision and the dissertation deadline; the
reliance on university participant pools with lower-than-anticipated response
rates; and the exclusion criterion (no prior GCSE Geography) which eliminated
a substantial proportion of Hull Computer Science students who had studied
Geography at school.

These shortcomings are disclosed here in the spirit of honest research
reporting, following the recommendation of \textcite{woolf2010} that ITS
evaluation papers should explicitly document the gap between planned and
achieved validation evidence.

% ─────────────────────────────────────────────────────────────────────────────
\section{Future Work}
\label{sec:disc_future}
% ─────────────────────────────────────────────────────────────────────────────

The following extensions are identified in priority order for continued
development:

\begin{enumerate}[leftmargin=2em]

  \item \textbf{Powered evaluation study.} Recruit $n \geq 18$ participants
    with a completed adaptive/static control condition and 7-day delayed
    retention assessment. This single study would transform the current
    proof-of-concept findings into publishable evidence and enable direct
    comparison with VanLehn's benchmark.

  \item \textbf{mirt recalibration.} Rerun IRT calibration using data from
    the evaluation sessions to produce more stable item parameter estimates.
    Monitor RMSEA and item fit statistics; flag items with poor fit for
    revision or removal.

  \item \textbf{Cross-Domain Knowledge Transfer.} Extend the KC graph to
    a second domain (e.g., GCSE History or Computer Science) using the
    existing domain-agnostic architecture. Implement inter-domain prerequisite
    edges to test whether learning in Geography KCs transfers to structurally
    related KCs in the second domain, following the KLI framework of
    \textcite{koedinger2012} and the cross-domain transfer model of
    \textcite{singley1989}.

  \item \textbf{Step-level feedback upgrade.} Redesign the feedback
    mechanism from post-hoc explanation (answer-based) to guided reasoning
    disclosure (step-based), bringing the system into alignment with
    VanLehn's high-effect-size category. This would require authoring
    multi-step solution traces per question item.

  \item \textbf{Longitudinal multi-session study.} Design a three-session
    protocol (baseline, adaptive, delayed retention) to evaluate the BKT
    decay detection mechanism and the successive relearning boost empirically.
    This is the only study design that can validate the cross-session BKT
    state persistence features, which remain unevaluated.

  \item \textbf{Secondary school deployment.} Pilot the system with UK KS3
    students in partnership with a geography department, providing
    ecologically valid engagement data with the intended population and
    enabling convergent validity assessment against GCSE practice paper
    performance.

\end{enumerate}

% ─────────────────────────────────────────────────────────────────────────────
\section{CSSE Reflection}
\label{sec:disc_csse}
% ─────────────────────────────────────────────────────────────────────────────

The CSSE competency requires demonstration of the ability to ``use established
software development methodologies to discover authentic requirements, respond
to change, ensure quality and deliver value to users and stakeholders.'' Five
episodes from this project's development are identified as direct evidence of
this competency.

\textbf{Discovering authentic requirements.} The initial system design was
derived from literature review rather than user research. Sprint 6 introduced
a first-visit onboarding tour following informal feedback from pilot users
who reported confusion about why the adaptive engine was presenting a
particular question. This is a concrete example of requirements discovery
through user contact rather than assumption: the transparency requirement was
not in the original specification, and was identified only through observed
user behaviour.

\textbf{Responding to change.} The Sprint 5 scope reduction from multi-domain
to single-domain validation was a managed response to emerging technical risk.
The decision involved: (a) identifying the risk explicitly (calibration data
infeasibility); (b) evaluating options (multi-domain with coarser calibration
vs.\ single-domain with rigorous calibration); (c) selecting the option that
better served the research goal; and (d) updating the project plan and Notion
tracker to reflect the new scope. This is textbook agile risk response, not
ad hoc retreat.

\textbf{Ensuring quality.} Three quality assurance mechanisms were
implemented: TypeScript strict mode eliminated a class of runtime errors;
Prisma migration files provide a versioned, auditable schema history; and
Vercel preview deployments enabled integration testing per feature branch
before merge. The Review Material bug (Section~\ref{subsec:impl_review_bug})
was identified during quality testing and documented honestly rather than
concealed --- demonstrating that quality assurance is a process of finding
and disclosing defects, not of claiming defect-free code.

A further quality episode arose during the write-up phase. The EAP
$\hat{\theta}$ estimation initially used a 21-point quadrature grid (step
size 0.3), which produced stepped ability trajectories rather than smooth
convergence. The error was identified by inspecting the algorithm validation
output and tracing the discontinuity to the grid resolution; extending to a
161-point grid resolved it. This illustrates a recurrent pattern in
psychometric software engineering: clean mathematical formulations do not
survive translation to fixed-precision computation unchanged, and validation
of numerical outputs against expected analytical behaviour is a distinct and
necessary quality assurance step beyond type checking and compilation.

\textbf{Auditing claims against implementation.} A late-stage cross-check
of dissertation claims against the live codebase revealed that the condition
assignment mechanism, while correctly specified in the database schema, was
not wired through to the signup API route
(Section~\ref{subsec:disc_condition_gap}). The discovery was made by asking
the question: ``does the implementation actually do what the report claims?''
and reading the relevant source file. This form of claim-against-code audit
is underused in student project development and is recommended as a
standard pre-submission practice. The fix was applied before final
submission; the evaluation data could not be retracted.

\textbf{Delivering value.} The system is production-deployed and accessible
to any learner with an internet connection. The open-source repository provides
value to other researchers and developers building adaptive systems for
non-STEM domains. The Misconceptions tab delivers actionable diagnostic
information that learners cannot obtain from a conventional quiz platform.
Each of these represents value delivered to a distinct stakeholder group
(learners, researchers, developers), demonstrating awareness that value
delivery is multi-stakeholder.
