\chapter{Conclusion}
\label{ch:conclusion}

% ─────────────────────────────────────────────────────────────────────────────
\section{Summary of Contributions}
\label{sec:conc_contributions}
% ─────────────────────────────────────────────────────────────────────────────

This project set out to design, implement, and evaluate a hybrid IRT 3PL +
BKT adaptive intelligent tutoring system for UK KS3/GCSE Geography. Five
concrete contributions have been made.

\textbf{First}, a production-deployed adaptive ITS for UK Geography was
built and made publicly accessible, filling a gap in a domain that has
received almost no attention in the adaptive learning literature
\parencite{bednarz2013}. The system is not a prototype or a simulation: it
handles real user accounts, persists learner state across sessions, and serves
live question content to any learner with an internet connection.

\textbf{Second}, a novel credit-factor modification to the standard BKT
learning transition equation was designed and implemented. This modification
formalises the evidential relationship between scaffolding usage and mastery
evidence by weighting the BKT posterior update according to the maximum hint
level accessed per question. To the author's knowledge, this specific
implementation (interpolating between the full Bayesian posterior and the
prior in proportion to the credit factor) has not been published in the
existing BKT literature, and represents a modest but concrete algorithmic
contribution.

\textbf{Third}, Bloom's taxonomy was implemented as a computational selection
constraint rather than a difficulty label. The Bloom-gated selection algorithm
enforces that learners consolidate recall-level knowledge before encountering
understanding and application items, implementing the cognitive scaffolding
principle of \textcite{anderson2001} as a live adaptive decision at each
question request.

\textbf{Fourth}, a five-track multi-method evaluation framework was developed
for small-sample ITS validation. This framework is reusable: any ITS project
facing the common constraint of insufficient sample size for inferential
statistics can adopt the same combination of algorithmic validation, engagement
data, usability metrics, misconception analysis, and robustness monitoring to
provide multi-strand evidence of system validity.

\textbf{Fifth}, a domain-agnostic open-source ITS architecture was produced.
The adaptive engine is parameterised entirely by KC identifiers and IRT
parameter files; it contains no Geography-specific logic. The codebase is
publicly available and provides a reusable implementation foundation for
researchers and developers building adaptive systems for declarative-knowledge
non-STEM domains.

% ─────────────────────────────────────────────────────────────────────────────
\section{Research Question Revisited}
\label{sec:conc_rq}
% ─────────────────────────────────────────────────────────────────────────────

The research question asked whether a hybrid IRT 3PL + BKT adaptive engine,
implemented in a production web application, can deliver personalised question
sequencing that produces measurable within-session learning gains for UK
KS3/GCSE Geography learners.

The answer is: \textit{yes, with qualification}. The algorithmic and technical
components of the claim are fully established: the engine selects items by
Fisher information at the learner's current $\hat{\theta}$, enforces
Bloom-level gates, tracks per-KC BKT state, and applies credit-factor
modifications for hint-assisted responses, all as specified, all in a
production deployment. The empirical component of the claim is supported at
proof-of-concept level by the $n = 4$ evaluation study, but cannot be
confirmed at statistical confidence without a powered study with a completed
control condition. This is an honest characterisation of the evidence, and
it is the appropriate characterisation: overclaiming empirical evidence is the
most common failure mode in ITS evaluation research \parencite{woolf2010}.

% ─────────────────────────────────────────────────────────────────────────────
\section{Broader Significance}
\label{sec:conc_significance}
% ─────────────────────────────────────────────────────────────────────────────

The broader significance of this project extends in three directions.

\textbf{ITS for non-STEM domains.} The preponderance of ITS research targets
mathematics and programming. This project demonstrates that the IRT+BKT
hybrid architecture transfers to a declarative-knowledge domain with a
multi-dimensional prerequisite structure, and that the pedagogical principles
of ZPD targeting, mastery gating, and spaced relearning are applicable to
Geography content. This extends the empirical and conceptual scope of ITS
research into territory that has been largely neglected.

\textbf{Open-source adaptive learning infrastructure.} Commercial adaptive
learning platforms (Khan Academy, DuoLingo, Knewton) are black boxes: their
algorithms are proprietary, their calibration data is unavailable for
independent scrutiny, and their architectures cannot be extended by
researchers. This project provides a fully transparent, auditable, and
extensible alternative. Every adaptive decision made by the engine can be
traced to specific parameter values in the open-source codebase. This
transparency matters for both academic scrutiny and learner trust.

\textbf{Domain-agnostic architecture as a research platform.} The system's
KC graph, IRT parameter store, and BKT state management are designed for
extension. The most significant future direction, Cross-Domain Knowledge
Transfer (in which learning in Geography KCs informs the prior estimates for
structurally related KCs in Mathematics or Computer Science), is architecturally
feasible without algorithmic changes. This positions the project not as a
finished product but as a research platform for investigating transfer learning
in adaptive educational systems.

% ─────────────────────────────────────────────────────────────────────────────
\section{Personal Reflection}
\label{sec:conc_reflection}
% ─────────────────────────────────────────────────────────────────────────────

This project required operating simultaneously across three domains of
expertise that are rarely combined: psychometrics (IRT calibration, EAP
estimation, BKT parameterisation), software engineering (Next.js architecture,
Prisma schema design, Vercel deployment, security hardening), and educational
research (study design, SUS administration, misconception analysis). The
integration of these domains was the project's primary intellectual challenge
and its primary source of value.

The scope reduction at Sprint 5 was the most professionally significant
decision of the project. The original ambition was larger; the final system
is more rigorous. That trade-off, depth over breadth, defensible claims over
impressive-sounding ones, is the central lesson of the project and one that
applies beyond academic work. A system that makes three well-evidenced claims
is more valuable than one that makes ten weakly-evidenced ones.

The experience of implementing a psychometric algorithm in a production web
application (handling the gap between the clean mathematics of the IRT
equations and the messy realities of JSON serialisation, race conditions, and
browser rendering) has provided a level of engineering judgement that no
coursework exercise could replicate. The Review Material bug, the EAP grid
resolution issue, the K-means supersession: these are not failures. They are
the data points from which engineering judgement is built.

This project has contributed to a professional portfolio spanning three
intersecting disciplines: psychometrics (IRT calibration, EAP estimation, BKT
parameterisation), software engineering (full-stack Next.js architecture,
production deployment, security hardening), and educational research (study
design, SUS administration, misconception analysis). Each of these disciplines
represents transferable knowledge that extends beyond the honours programme and
contributes to a foundation for lifelong learning in educational technology,
data science, and software architecture. The decision to work across domain
boundaries, rather than optimising depth in a single area, reflects a
deliberate professional development strategy: the highest-value engineering
problems in the next decade will consistently sit at the intersection of
technical implementation and domain expertise.

% ─────────────────────────────────────────────────────────────────────────────
\section{Final Statement}
\label{sec:conc_final}
% ─────────────────────────────────────────────────────────────────────────────

Bloom's 2-sigma problem has not been solved. Human tutoring remains superior
to any automated system across the full range of educational contexts and
learner profiles. But the problem is tractable: the literature shows that
well-designed adaptive systems can close the gap substantially, and that the
key design variables are interaction granularity, principled algorithm
selection, and transparent feedback, not algorithmic complexity or scale.

GeoMentor is a small contribution to a large problem. It demonstrates that
these design principles can be implemented at production quality by a single
developer, in a non-STEM domain, using open-source tools, within an academic
project timeline. If it serves as a foundation for a powered evaluation study,
a secondary school pilot, or a researcher building an adaptive system for
History or Computer Science, it will have served its purpose.
